---
title: "All Service Center Reports"
author: "Selenea Gibson & Jason Howard"
date: "2025-01-02"
output: html_document
---

# Project Overview
This markdown file will use the DHCD Finance dept documents that are for reporting expenditures and GL data. The data is collected from Workday. This file includes all of the service centers and there cost centers in the rmd. 

```{r libraries, warning=FALSE, message=FALSE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
```

## Reading in the data 
Reading in the files for the data that will be used. Put new data files in the "Data" folder and the script will automatically detect the latest file within the Data folder.

```{r data}
get_latest_file <- function(pattern, date_format = "%Y-%m-%d") {
  files <- list.files(path = "Data", pattern = pattern, full.names = TRUE)
  
  if (length(files) == 0) return(NULL)  # Return NULL if no matching files are found
  
  # Extract dates using regex that matches "YYYY-M-D"
  file_dates <- as.Date(sub(".*?(\\d{4}-\\d{1,2}-\\d{1,2}).*", "\\1", files), format = date_format)
  
  # Get the file with the latest date
  latest_file <- files[which.max(file_dates)]
  
  return(latest_file)
}

# assigning the names of the latest files to the new variables
gl_all_data <- basename(get_latest_file("GL-Operating Actual Expenses YTD.*\\.xlsx$"))
budg_vs_act_file_name <- basename(get_latest_file("Budget_vs_Actuals_-_BBMR YTD.*\\.xlsx$"))

# read data from files
all_gl_data <- readxl::read_xlsx(here("Data", gl_all_data), 
                             skip = 14,
                             col_types = c(rep("text",2),   
                                           "date", 
                                           rep("text",6),
                                           rep("numeric", 3),
                                           rep("text",26))) |> 
  mutate(`Accounting Date` = as.Date(`Accounting Date`))

# the first row of the data file is mostly empty except for a few important dates. 
# read those data and do some formatting to clean up the data
first_row_of_data <- 
  readxl::read_xlsx(here("Data", budg_vs_act_file_name)) |> 
  names()

# take out the data that is before the numeric values
name_prefixes <- 
  tibble(first_row_of_data) |> 
  mutate(prefixes = ifelse(str_detect(first_row_of_data, "\\.\\."), 
                           NA_character_, 
                           first_row_of_data)) |> 
  fill(prefixes, .direction = "down") |> 
  mutate(prefixes = ifelse(is.na(prefixes), 
                           NA_character_, 
                           paste0(prefixes, "-01"))) |> 
  pull(prefixes)

# read the data file again but skip the first line
budg_vs_actuals <- 
  readxl::read_xlsx(here("Data", budg_vs_act_file_name),
                    skip = 1)

# remove the total budget row as this is not needed 
# and will mess up the data. 
budg_vs_actuals <- budg_vs_actuals |> 
  filter(`Cost Center` != "Total") 

# the names in the file repeat themselves, but this combines the dates with the repetitive names
for(i in seq_along(name_prefixes)){
  plain_name <- gsub("\\.\\..*","", names(budg_vs_actuals)[i]) 
  
  names(budg_vs_actuals)[i] <- 
    ifelse(is.na(name_prefixes[i]) & str_length(plain_name) > 0,
           plain_name,
           paste(plain_name,name_prefixes[i], sep = "_"))
}

# remove commitments and obligations since we dont need them
budg_vs_actuals <- budg_vs_actuals |> 
  select(-matches("^Commitments|^Obligations"))

# looking at the new names 
names(budg_vs_actuals)
```


## Cleaning the datasets
I need to remove some of the rows that are not needed for the analysis. I will then make the new row be the header row. 
We need to remove the accruals in the GL dataset as those are not needed to be in the file. These are known as (Reversal) in the dataset.

```{r cleaningdata}
# add round all expenditure days to the expenditure month
all_gl_data <- all_gl_data |> 
  mutate(expend_month = floor_date(`Accounting Date` , unit = "month"))

# remove columns that have nulls for all values 
# replace "blank" with an actual 'null' value
budg_vs_actuals <- budg_vs_actuals |> 
  select_if(~sum(!is.na(.)) > 0) |> 
  mutate(Grant = ifelse(Grant == "(Blank)", NA_character_, Grant))
```


####
## Creating the fiscal months
This will create the fiscal months and year that can be used for `754 SFSP Expense Report`. 

```{r, warning=FALSE}
# -- identifying the months of the fiscal year ---
# find all months of fiscal year
all_column_names <- names(budg_vs_actuals)
month_columns <- all_column_names[str_detect(all_column_names, "Actuals_")]
month_years <- str_extract(month_columns, "(?<=Actuals_)(\\d{4}-\\d{2})")  # Extracts 'YYYY-MM'
months <- as.numeric(substr(month_years, 6, 7)) 
fiscal_months <- ifelse(months >= 8, months - 7, months + 5)
current_month <- month(Sys.Date())  # Get the current month (e.g., February = 2)
latest_fiscal_month <- ifelse(current_month >= 8, current_month - 7, current_month + 5)
# calculate the previous fiscal month
month_dates <- as.Date(paste0(month_years, "-01"))
get_first_date <- min(month_dates, na.rm = TRUE)
all_year_months <- seq(get_first_date, get_first_date + months(11), by = "month")

# --- placeholders for the missing months ---
# add missing months into the table
for (col in paste0("Actuals_", all_year_months)) {
  if (!col %in% names(budg_vs_actuals)) {
    budg_vs_actuals[[col]] <- NA_real_  # Add missing columns filled with NA
  }
}

# Build the expense report table dynamically
all_expense_report_tbl <- budg_vs_actuals |> 
  group_by(Fund, `Spend Category`, `Cost Center`, Service) |>  # Now including Service
  summarise(
    across(where(is.numeric), ~ sum(.x, na.rm = TRUE)),  # Sum Actuals columns
    .groups = "drop"
  )


# Rename columns to the "Month Year" format (e.g., "June 2024")
all_expense_report_tbl <- all_expense_report_tbl |> 
  rename_with(
    ~ format(as.Date(gsub("Actuals_", "", .x)), "%B %Y"),
    starts_with("Actuals_")
  )

```

###
## Calculations for gl_ytd_exp
Calculating the gl_ytd_exp column. I will also rename the `Total Actual` column to the `BBMR Total Expenditure` column. Then join the gl table to the all expense report table.

```{r gl}
# --- aggregate the GL_YTD_EXP by Fund, Spend Category, and Cost Center ---
all_gl_data_exp <- all_gl_data |> 
  group_by(Fund, `Spend Category as Worktag`, `Cost Center`, `Service (Cost Center Hierarchy)`) |> 
  summarise(`GL YTD EXP` = sum(Amount, na.rm = TRUE), .groups = "drop")

# --- Joining datasets ---
# joining the gl YTD data to the all_expense_report_tbl
gl_final_report <- all_expense_report_tbl |> 
  left_join(all_gl_data_exp, 
            by = c("Spend Category" = "Spend Category as Worktag", 
                   "Cost Center" = "Cost Center", 
                   "Fund" = "Fund")) |> 
  rename(`Total Budget` = `Revised Budget`) |> 
  mutate(`GL YTD EXP` = if_else(is.na(`GL YTD EXP`), 0, `GL YTD EXP`))

# arrange and drop uneeded columns
gl_final_report <- gl_final_report |> 
  select(-c("Service (Cost Center Hierarchy)", 
            "Amendments",
            "Corrections to Budget",
            "Total Actuals",
            "Total Spent",
            "Adopted Budget")) |> 
select(Service, Fund, `Cost Center`, `Spend Category`, everything())  # Reorders columns
```

  
###
## Calculations for BBMR_ytd_exp
Calculating the BBMR_ytd_exp column. Then join the BBMR table to the all expense report table. 

```{r bbmr, warning=FALSE, message=FALSE}
# --- aggregate the BBMR YTD EXP by Fund, Spend Category, and Cost Center ---

# --- Extract the month columns ---
month_cols <- names(budg_vs_actuals)[15:26]  # Adjust if necessary

# --- Summing monthly values across Cost Centers and Service ---
all_bbmr_data_exp_total <- budg_vs_actuals |> 
  group_by(`Service`, Fund, `Cost Center`) |>  # Removed Spend Category
  summarise(across(all_of(month_cols), sum, na.rm = TRUE), .groups = "drop") 

# Ensure month_cols are numeric before rowSums()
all_bbmr_data_exp_total <- all_bbmr_data_exp_total |> 
  mutate(`BBMR YTD EXP` = rowSums(across(all_of(month_cols)), na.rm = TRUE))

# --- Summing the Revised Budget for each Cost Center ---
revised_budget_summary <- all_expense_report_tbl |> 
  group_by(`Cost Center`) |> 
  summarise(`Total Budget` = sum(`Revised Budget`, na.rm = TRUE), .groups = "drop")

expense_unique <- all_expense_report_tbl |> 
  select(Fund, `Cost Center`, `Service`) |>  # Removed Spend Category
  distinct() |>  # Remove duplicates
  filter(!is.na(`Service`))  # Remove NAs in Service

# --- Merging all aggregated data into one final dataset ---
bbmr_final_data <- all_bbmr_data_exp_total |> 
  left_join(expense_unique, by = c("Cost Center", "Service", "Fund")) |>  # Ensuring correct join
  left_join(revised_budget_summary, by = "Cost Center")  # Merge Total Budget

# need to make the months be the correct names and not like the weird names
bbmr_final_data <- bbmr_final_data |> 
  rename_with(
  ~ as.character(month(as.Date(gsub("Actuals_","", .x)), label = T, abbr = F)),
  starts_with("Actuals_")
)

# lastly need to make these column allign with the main report
bbmr_final_report <- bbmr_final_data |> 
  arrange(Service, Fund, `Cost Center`, `Total Budget`) |>  # Sorts rows correctly
  select(Service, Fund, `Cost Center`, `Total Budget`, everything())  # Reorders columns
```

####
### Creating the Excel workbooks
####

###
## Creating the workbooks for Excel 
Creating 3 workbooks that will contain the data from the BBMR and GL report. This will be used to create the final product for the `754 SFSP Expense Report` table. 

1. Calculating the columns for the Analysis Table - GL
Create the columns that are needed for the second table that is in the `754 SFSP Expense Report`.
*NOTE: This is for the GL only*


```{r calculatingcols, warning=FALSE}
# create a new dataframe with only the new columns included and not the rest
# all_tblwhodis is a placeholder for the new calculated columns

# --- calculate the Remaining Balance column ---

# create a new column for the remaining balance and make the gl ytd exp numeric
gl_all_tblwhodis <- gl_final_report |> 
  mutate(
    `Total Budget` = as.numeric(gsub("[^0-9.-]", "", `Total Budget`)),
    `GL YTD EXP` = as.numeric(gsub("[^0-9.-]", "", `GL YTD EXP`)),
    `Remaining Balance` = ifelse(
      is.na(`Total Budget`) | is.na(`GL YTD EXP`), 
      NA,  
      `Total Budget` - `GL YTD EXP`  
    )
  )

# --- Projected Y|E Value calculation ---

# calculating the projected Y|E Values
gl_all_tblwhodis <- gl_all_tblwhodis |> 
  mutate(
    FiscalMonth = latest_fiscal_month,  
    `Projected Y|E Value` = ifelse(
      is.na(`GL YTD EXP`) | is.na(FiscalMonth), 
      NA,  
      (`GL YTD EXP` / FiscalMonth) * 12  
    )
  ) |> 
  select(-FiscalMonth)  # Drop FiscalMonth column while keeping everything else

# --- Sur / (Def) column ---

# calculate the Sur / (Def) column 
gl_all_tblwhodis <- gl_all_tblwhodis |>
  mutate(
    `Sur / (Def)` = ifelse(
      is.na(`Total Budget`) | is.na(`Projected Y|E Value`),
      NA,  
      `Total Budget` - `Projected Y|E Value`  
    )
  )


# --- Notes column ---# ---`Projected Y|E Value` Notes column ---

# here we will make a notes column that can be used in excel
# initialize as empty character column
gl_all_tblwhodis$Notes <- NA_character_  

# save this file for gl plotting later
writexl::write_xlsx(gl_all_tblwhodis, "gl_all_tblwhodis.xlsx")
```


###
## Create the Excel workbook for wb1 and wb2
Creating wb1 for the all service centers and wb2 for the analysis all table.

```{r creatingwbs}
# workbook 1 for gl 
# gl_final_report


# workbook 2 for gl analysis table
# only keep the columns that are needed for this table
gl_analysis_tbl <- gl_all_tblwhodis |> 
  select(`Remaining Balance`, `Projected Y|E Value`, `Sur / (Def)`, Notes)

# workbook 3 for bbmr 
# bbmr_final_report
```



###
## Styling for the Excel workbooks
This will create the styling for the 3 wbs. The All Service Centers Report GL and All Service Centers Report BBMR, and the Analysis Table for the GL. 
*NOTE: You do NOT need to change the name of the title header for the tables. As the function in line 284 will automatically detect the fiscal year end or begining date and apply the month/day to the header (line 294)*

*This is for the GL table only!*

```{r stylingforwb1}
# --- Get the current system date when this was run ---
latest_month <- all_gl_data |> 
  filter(!is.na(expend_month)) |>      # Ensure we only consider filled-out data
  summarize(latest_date = max(expend_month, na.rm = TRUE)) |> 
  pull(latest_date) |> 
  month()

latest_month_name <- month.name[latest_month]

# Extract year of the latest date
latest_year <- all_gl_data |> 
  filter(!is.na(expend_month)) |> 
  summarize(latest_date = max(expend_month, na.rm = TRUE)) |> 
  pull(latest_date) |> 
  year()

# Extract the current system date
current_date <- Sys.Date() - 1
current_year <- year(current_date)
current_month <- month(current_date) -1

previous_month_date <- current_date %m-% months(1)
month_name_forend <- month.name[month(previous_month_date)]

# Determine the fiscal year
if (current_month >= 7) {
  fiscal_year <- current_year + 1
} else {
  fiscal_year <- current_year
}

# --- Create a Workbook ---

# Format the title header 
title_header <- paste0("FY ", fiscal_year, " Budget and Actual as of ", month_name_forend, " ", latest_year)

# Create the workbook
wb1 <- createWorkbook()
addWorksheet(wb1, "All Service Centers Report GL")

# --- Add Title (Row 1) ---

# Write the header title
writeData(wb1, "All Service Centers Report GL", title_header, startRow = 1, startCol = 1)

# Define the title style
header_style <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#ED7D31")

# Merge cells across all columns for the title
mergeCells(wb1, "All Service Centers Report GL", cols = 1:ncol(gl_final_report), rows = 1)
addStyle(wb1, "All Service Centers Report GL", header_style, rows = 1, cols = 1:ncol(gl_final_report), gridExpand = TRUE)

# --- Write the Data (Row 2) with Column Names ---

writeData(wb1, "All Service Centers Report GL", gl_final_report, startRow = 2, startCol = 1, colNames = TRUE)

# --- Align and Style Column Headers ---

# Center alignment for column headers
center_style <- createStyle(halign = "center", valign = "center")
addStyle(wb1, "All Service Centers Report GL", center_style, rows = 2, cols = 1:ncol(gl_final_report), gridExpand = TRUE)

# Define column header style
column_names_style <- createStyle(
  halign = "center", 
  valign = "center", 
  textDecoration = "bold", 
  fgFill = "#ED7D31"
)

# Apply the style to the column names
addStyle(wb1, "All Service Centers Report GL", column_names_style, rows = 2, cols = 1:ncol(gl_final_report), gridExpand = TRUE)

# --- Save the Workbook ---

saveWorkbook(wb1, "all_service_center_report_GL.xlsx", overwrite = TRUE)
```

Styling for the 2 wb or Analysis all table.

```{r stylingforwb2}
# --- create a workbook ---

# add the worksheet to the created workbook
wb2 <- createWorkbook()
addWorksheet(wb2, "all_analysis_tbl")

# --- write data to the worksheet 2 ---

writeData(wb2, "all_analysis_tbl", gl_analysis_tbl, startRow = 2, startCol = 1, colNames = TRUE)

# --- header for the table---

# add a header, style, and color with the label "Analysis" in the first row
writeData(wb2, "all_analysis_tbl", "ANALYSIS", startRow = 1, startCol = 1)
header_style <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4")

# merge the cells
mergeCells(wb2, "all_analysis_tbl", cols = 1:ncol(gl_analysis_tbl), rows = 1)
addStyle(wb2, "all_analysis_tbl", header_style, rows = 1, cols = 1:ncol(gl_analysis_tbl), gridExpand = TRUE)

# --- align for column headers ---

# center alignment for column headers
center_style <- createStyle(halign = "center", valign = "center")
addStyle(wb2, "all_analysis_tbl", center_style, rows = 2, cols = 1:ncol(gl_analysis_tbl), gridExpand = TRUE)

# --- style the column names (e.g., "Remaining Balance", "Sur / (Def)", etc.) ---

# styling the column names for the datatable
column_names_style <- createStyle(
  halign = "center", 
  valign = "center", 
  textDecoration = "bold", 
  fgFill = "#C6E0B4" 
)

# apply the style to the column names (first row of data)
addStyle(wb2, "all_analysis_tbl", column_names_style, rows = 2, cols = 1:ncol(gl_analysis_tbl), gridExpand = TRUE)

# save the workbook
saveWorkbook(wb2, "all_analysis_tbl.xlsx", overwrite = TRUE)
```


Styling for the 3 wb or BBMR table.

```{r stylingforwb3}
# --- Create a Workbook ---

# Format the title header with detected month and fiscal year
title_header <- paste0("FY ", fiscal_year, " Budget and Actual as of ", month_name_forend, " ", latest_year)

# Create the workbook
wb3 <- createWorkbook()
addWorksheet(wb3, "All Service Centers Report BBMR")

# --- Add Title (Row 1) ---

# Write the header title
writeData(wb3, "All Service Centers Report BBMR", title_header, startRow = 1, startCol = 1)

# Define the title style
header_style <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#ED7D31")

# Merge cells across all columns for the title
mergeCells(wb3, "All Service Centers Report BBMR", cols = 1:ncol(bbmr_final_report), rows = 1)
addStyle(wb3, "All Service Centers Report BBMR", header_style, rows = 1, cols = 1:ncol(bbmr_final_report), gridExpand = TRUE)

# --- Write the Data (Row 2) with Column Names ---

writeData(wb3, "All Service Centers Report BBMR", bbmr_final_report, startRow = 2, startCol = 1, colNames = TRUE)

# --- Align and Style Column Headers ---

# Center alignment for column headers
center_style <- createStyle(halign = "center", valign = "center")
addStyle(wb3, "All Service Centers Report BBMR", center_style, rows = 2, cols = 1:ncol(bbmr_final_report), gridExpand = TRUE)

# Define column header style
column_names_style <- createStyle(
  halign = "center", 
  valign = "center", 
  textDecoration = "bold", 
  fgFill = "#ED7D31"
)

# Apply the style to the column names
addStyle(wb3, "All Service Centers Report BBMR", column_names_style, rows = 2, cols = 1:ncol(bbmr_final_report), gridExpand = TRUE)

# --- Save the Workbook ---

saveWorkbook(wb3, "all_service_center_report_BBMR.xlsx", overwrite = TRUE)

# saving the workbook
# the file will automatically assign a date to the workbook
saveWorkbook(wb3, paste0("FY ", fiscal_year, " BBMR All Services Budget and Actual as of ", month_name_forend, " ", latest_year, ".xlsx"), overwrite = TRUE)
```


###
## Combining the workbooks
Last step, "Whew!", is to join the two of these workbooks together to have 1 full working workbook that can be used for the team. 

```{r combiningwbs}
# --- combine the two workbooks on the same worksheet with full formatting ---

# create a new workbook for the combined data
all_wb_combined <- createWorkbook()

# add a worksheet for combined data
addWorksheet(all_wb_combined, "tables_unite")

# define styles
title_style_summer <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#FFA500", fontSize = 11)
title_style_analysis <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4", fontSize = 11)

header_style_summer <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#FFA500")
header_style_analysis <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4")

# --- write the first table (gl_final_report) with title and formatting ---

# add title for the first table
writeData(all_wb_combined, "tables_unite", paste("All Service Centers Report GL as of", month_name_forend), startRow = 1, startCol = 1)
mergeCells(all_wb_combined, "tables_unite", cols = 1:ncol(gl_final_report), rows = 1)
addStyle(all_wb_combined, "tables_unite", title_style_summer, rows = 1, cols = 1:ncol(gl_final_report), gridExpand = TRUE)

# write the first table
writeData(all_wb_combined, "tables_unite", gl_final_report, startRow = 2, startCol = 1, colNames = TRUE)

# apply column header style to the first table
addStyle(all_wb_combined, "tables_unite", header_style_summer, rows = 2, cols = 1:ncol(gl_final_report), gridExpand = TRUE)

# --- add 2 blank columns for separation ---

# count the number of rows and columns
num_rows_summer <- nrow(gl_final_report)
num_cols_summer <- ncol(gl_final_report)

# write blank columns for separation
writeData(all_wb_combined, "tables_unite", matrix("", nrow = num_rows_summer + 2, ncol = 2), 
          startRow = 1, startCol = num_cols_summer + 1, colNames = FALSE)

# --- Write the second table (gl_analysis_tbl) with title and formatting ---

# add title for the second table
writeData(all_wb_combined, "tables_unite", "Analysis Table", startRow = 1, startCol = num_cols_summer + 3)
mergeCells(all_wb_combined, "tables_unite", cols = (num_cols_summer + 3):(num_cols_summer + 2 + ncol(gl_analysis_tbl)), rows = 1)
addStyle(all_wb_combined, "tables_unite", title_style_analysis, rows = 1, cols = (num_cols_summer + 3):(num_cols_summer + 2 + ncol(gl_analysis_tbl)), gridExpand = TRUE)

# write the second table
writeData(all_wb_combined, "tables_unite", gl_analysis_tbl, startRow = 2, startCol = num_cols_summer + 3, colNames = TRUE)

# apply column header style to the second table
addStyle(all_wb_combined, "tables_unite", header_style_analysis, rows = 2, cols = (num_cols_summer + 3):(num_cols_summer + 2 + ncol(gl_analysis_tbl)), gridExpand = TRUE)

# --- save the combined workbook ---

# saving the workbook
# the file will automatically assign a date to the workbook
saveWorkbook(all_wb_combined, paste0("FY ", fiscal_year, " All Services Budget and Actual as of ", month_name_forend, " ", latest_year, ".xlsx"), overwrite = TRUE)
```

#### YOU DID IT ####
Congratulations! You did a wonderful job coding! 
If this is your first real coding script, then I must say you choose a tricky script. 
Go ahead and call yourself a Rstudio programmer you data analyst!
#### YOU DID IT ####












