---
title: "All Service Centers Plots"
author: "Selenea Gibson and Jason Howard"
date: "2025-01-23"
output: html_document
---

# Project Overview
This markdown file will use the DHCD Finance dept data to calculate the columns that are provided from the `754 SFSP Expense Report`. There will be a series of plots to make the tables at the bottom of the report more appealing for what is being shown. This will be done through visuals.

*Note: This rmd file contains all of the cost centers not just the summer food program. The numbers will be a lot larger than the summer food program plots file.*

```{r libraries, warning=FALSE, message=TRUE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
library(hrbrthemes)
library(ggplot2)
```


## Reading in the data 
Reading in the files for the data that will be used for creating the plots that will represent the data on the bottom of the `754 SFSP Expense Report`.

```{r readingdata}
# reading in the dataset for the summer_expense_report
allbud_vs_actuals_files <- list.files(here(), pattern = "FY \\d{4} All Services Budget and Actual.*\\.xlsx", full.names = TRUE)

# Read the latest file
allbud_vs_actuals <- read.xlsx(allbud_vs_actuals_files)

# set the first row as headers and remove it from the data
allbud_vs_actuals <- setNames(allbud_vs_actuals[-1, ], as.character(unlist(allbud_vs_actuals[1, ])))  
```


## Sorting the data to contain only the 1001 General and 5000 State Grants Fund data.
This will be a dataset that contains the 1001 General Fund and the 5000 State Grant Funds. This will be used for the analysis below then compared with each of the Funds.


```{r allgfsorted}
# sort through the general and state fund
all_general_fund <- allbud_vs_actuals |> filter(Fund == '1001 General Fund')
all_state_fund <- allbud_vs_actuals |> filter(Fund == '5000 State Grants Fund')
```


####
## Creating calculations for the tables
Creating the calculations for the tables. There will be a total of 10 tables that are from the `754 SFSP Expense Report`. 

*First calculations include the Revised Budget*
*Second calculations include the GL_YTD_EXP*
*Third calculations include the Analysis Table*
*Fourth calculation include the Percentage Table*
####


### General Funds 1001 Table
*First calculations include the Revised Budget*
Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 1001 General Funds.

```{r gfloop}
# --- define salary-related keywords ---

salary_keywords <- c(
  "Permanent Full-Time", "FICA Payroll Tax", "Overtime", "Permission Time",
  "Employees' Retirement System \\(ERS\\) Pension", "Medical Insurance",
  "Prescription Drug Insurance", "Vision Insurance", "Survivor Benefits",
  "Dental Insurance", "Budgeted FICA Payroll Tax \\(Temporary Employees\\)", 
  "Comp Time", "Meal Allowance", "Temporary", "Budgeted Pending Personnel Actions"
)

# --- create regex pattern for salary-related terms ---

salary_pattern <- paste(salary_keywords, collapse = "|")

# --- extract unique cost centers from General Fund dataset ---

gf_cost_center_list <- unique(all_general_fund$`Cost Center`)

# --- initialize list to store results ---

general_fund_results <- list()

# --- loop through each Cost Center ---

for (cost_center in gf_cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_general_fund |> filter(`Cost Center` == cost_center)
  
  # Compute salary and non-salary projections
  gf_salary_proj <- sum(as.numeric(cost_center_data$`Revised Budget`[str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  gf_non_salary_proj <- sum(as.numeric(cost_center_data$`Revised Budget`[!str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  
  # Store results in a dataframe
  general_fund_results[[as.character(cost_center)]] <- data.frame(
    `Cost Center` = cost_center,
    `Salaries/OPC Projections` = gf_salary_proj,
    `Other EXP Projections` = gf_non_salary_proj,  # Ensured gf_ prefix
    `Total Projections` = gf_salary_proj + gf_non_salary_proj  # Consistent naming
  )
}

# --- combine all cost centers into a final dataframe ---

general_fund_summary_df <- bind_rows(general_fund_results)
```


### Other Funds - 2089, 4000, and 5000 table
Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 5000 State Grant Funds.


```{r sgfunction}
# --- extract unique cost centers from State Fund dataset ---

sf_cost_center_list <- unique(all_state_fund$`Cost Center`)

# --- initialize list to store results ---

state_fund_results <- list()

# --- loop through each Cost Center ---

for (cost_center in sf_cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_state_fund |> filter(`Cost Center` == cost_center)
  
  # Compute salary and non-salary projections
  sf_salary_proj <- sum(as.numeric(cost_center_data$`Revised Budget`[str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  sf_non_salary_proj <- sum(as.numeric(cost_center_data$`Revised Budget`[!str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  
  # Store results in a dataframe
  state_fund_results[[as.character(cost_center)]] <- data.frame(
    `Cost Center` = cost_center,
    `Salaries/OPC Projections` = sf_salary_proj,
    `Other EXP Projections` = sf_non_salary_proj,  
    `Total Projections` = sf_salary_proj + sf_non_salary_proj  
  )
}

# --- combine all cost centers into a final dataframe ---

state_fund_summary_df <- bind_rows(state_fund_results)
```


### Total Projections equals to row 268 
Combining the general and state funds for the tpe table. 

```{r tpetblone, warning=FALSE}
# --- Bind the rows together ---
tpe_summary <- bind_rows(general_fund_summary_df, state_fund_summary_df)

# --- Create the tpe_rb_final_tbl table ---
tpe_rb_final_tbl <- tpe_summary |> 
  summarise(
    Salaries.OPC.Projections = sum(as.numeric(Salaries.OPC.Projections), na.rm = TRUE),
    Other.EXP.Projections = sum(as.numeric(Other.EXP.Projections), na.rm = TRUE),
    Total.Projections = Salaries.OPC.Projections + Other.EXP.Projections
  )
```



### GL_YTD_EXP Tables
*Second calculations include the GL_YTD_EXP*

## GL_YTD_EXP General Funds 1001 Table
Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 1001 General Funds.

```{r gfglyplot}
# --- initialize list to store results ---
glytd_general_fund_results <- list()

# --- loop through each Cost Center ---
for (cost_center in gf_cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_general_fund |> filter(`Cost Center` == cost_center)
  
  # Compute salary and non-salary projections
  glytd_salary_proj <- sum(as.numeric(cost_center_data$`GL YTD EXP`[str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  glytd_non_salary_proj <- sum(as.numeric(cost_center_data$`GL YTD EXP`[!str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  
  # Store results in a dataframe and save in the list
  glytd_general_fund_results[[as.character(cost_center)]] <- data.frame(
    `Cost Center` = cost_center,
    `Salaries/OPC Projections` = glytd_salary_proj,
    `Other EXP Projections` = glytd_non_salary_proj,
    `Total Projections` = glytd_salary_proj + glytd_non_salary_proj
  )
}


# --- combine all cost centers into a final dataframe ---

glytd_general_fund_summary_df <- bind_rows(glytd_general_fund_results)
```


### Other Funds - 2089, 4000, and 5000 table
Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 5000 State Grant Funds.

```{r ofgl, warning=FALSE}
# --- initialize list to store results ---

glytd_state_fund_results <- list()

# --- loop through each Cost Center ---

for (cost_center in sf_cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_state_fund |> filter(`Cost Center` == cost_center)
  
  # Compute salary and non-salary projections
  glytd_sf_salary_proj <- sum(as.numeric(cost_center_data$`GL YTD EXP`[str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  glytd_sf_non_salary_proj <- sum(as.numeric(cost_center_data$`GL YTD EXP`[!str_detect(cost_center_data$`Spend Category`, salary_pattern)]), na.rm = TRUE)
  
  # Store results in a dataframe
  glytd_state_fund_results[[as.character(cost_center)]] <- data.frame(
    `Cost Center` = cost_center,
    `Salaries/OPC Projections` = glytd_sf_salary_proj,
    `Other EXP Projections` = glytd_sf_non_salary_proj,
    `Total Projections` = glytd_sf_salary_proj + glytd_sf_non_salary_proj
  )
}

# --- combine all cost centers into a final dataframe ---

glytd_state_fund_summary_df <- bind_rows(glytd_state_fund_results)
```


### Total Projections equals to row 268 
This is for the GL_YTD_EXP Tables. Combining the general and state funds for the tpe table. 

```{r tpetwo, warning=FALSE}
# --- bind the rows together ---

glytd_tpe_summary <- bind_rows(glytd_general_fund_summary_df, glytd_state_fund_summary_df) 

# --- create the final TPE summary table for plotting ---

glytd_tpe_final_tbl <- glytd_tpe_summary |> 
  summarise(
    Salaries.OPC.Projections = sum(as.numeric(Salaries.OPC.Projections), na.rm = TRUE),
    Other.EXP.Projections = sum(as.numeric(Other.EXP.Projections), na.rm = TRUE),
    Total.Projections = Salaries.OPC.Projections + Other.EXP.Projections
  )
```


*Third calculations include the Analysis Table*

### General Funds 1001 Table
This is for the Analysis Tables. Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 1001 General Funds.

```{r antblone, warning=FALSE}
# --- Initialize list to store results ---

alys_general_fund_results <- list()

# --- Loop through each Cost Center ---

cost_center_list <- unique(all_general_fund$`Cost Center`)

for (cost_center in cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_general_fund |> filter(`Cost Center` == cost_center)
  
  # Separate salary-related rows and non-salary-related rows
  salary_rows <- cost_center_data |> 
    filter(str_detect(`Spend Category`, salary_pattern))

  non_salary_rows <- cost_center_data |> 
    filter(!str_detect(`Spend Category`, salary_pattern))

  # Calculate Salaries/OPC projections
  salary_proj_remaining_balance <- sum(as.numeric(all_general_fund$`Remaining Balance`), na.rm = TRUE)
  salary_proj_projected_value <- sum(as.numeric(all_general_fund$`Projected Y|E Value`), na.rm = TRUE)
  salary_proj_sur_def <- sum(as.numeric(all_general_fund$`Sur / (Def)`), na.rm = TRUE)
  
  # Calculate Other EXP projections
  non_salary_proj_remaining_balance <- sum(as.numeric(all_general_fund$`Remaining Balance`), na.rm = TRUE)
  non_salary_proj_projected_value <- sum(as.numeric(all_general_fund$`Projected Y|E Value`), na.rm = TRUE)
  non_salary_proj_sur_def <- sum(as.numeric(all_general_fund$`Sur / (Def)`), na.rm = TRUE)
  
  # Store results for this cost center in a data frame
  alys_gf_summary <- data.frame(
    `Cost Center` = cost_center,
    `Remaining Balance - Salaries/OPC` = salary_proj_remaining_balance,
    `Projected YE Value - Salaries/OPC` = salary_proj_projected_value,
    `Sur/(Def) - Salaries/OPC` = salary_proj_sur_def,
    `Remaining Balance - Other EXP` = non_salary_proj_remaining_balance,
    `Projected YE Value - Other EXP` = non_salary_proj_projected_value,
    `Sur/(Def) - Other EXP` = non_salary_proj_sur_def
  )
  
  # Store this cost center summary in the results list
  alys_gf_final_tbl <- alys_gf_summary |> 
    summarise(
      `Salaries/OPC Projections` = sum(`Remaining.Balance...Salaries.OPC`, na.rm = TRUE) +
                                   sum(`Projected.YE.Value...Salaries.OPC`, na.rm = TRUE) +
                                   sum(`Sur..Def....Salaries.OPC`, na.rm = TRUE),
      `Other EXP Projections` = sum(`Remaining.Balance...Other.EXP`, na.rm = TRUE) +
                                sum(`Projected.YE.Value...Other.EXP`, na.rm = TRUE) +
                                sum(`Sur..Def....Other.EXP`, na.rm = TRUE),
      `Total Projections` = `Salaries/OPC Projections` + `Other EXP Projections`
    )
  
  # Append the final table to the results list
  alys_general_fund_results[[as.character(cost_center)]] <- alys_gf_final_tbl
}

# --- Combine all results into a final dataframe ---
alys_general_fund_summary_df <- bind_rows(alys_general_fund_results) |> 
  slice(1)
```


### Other Funds - 2089, 4000, and 5000 table
This is for the Analysis Tables. Here is a for loop that will automatically configure the key words that match up with salaries and non spend categories for only the 5000 State Grant Funds.

```{r anlysofloop, warning=FALSE}
# --- Initialize list to store results ---

alys_state_fund_results <- list()

# --- Loop through each Cost Center ---

cost_center_list <- unique(all_state_fund$`Cost Center`)

for (cost_center in cost_center_list) {
  # Filter data for the current cost center
  cost_center_data <- all_state_fund |> filter(`Cost Center` == cost_center)
  
  # Separate salary-related rows and non-salary-related rows
  salary_rows <- cost_center_data |> 
    filter(str_detect(`Spend Category`, salary_pattern))

  non_salary_rows <- cost_center_data |> 
    filter(!str_detect(`Spend Category`, salary_pattern))

  # Calculate Salaries/OPC projections
  salary_proj_remaining_balance <- sum(as.numeric(all_state_fund$`Remaining Balance`), na.rm = TRUE)
  salary_proj_projected_value <- sum(as.numeric(all_state_fund$`Projected Y|E Value`), na.rm = TRUE)
  salary_proj_sur_def <- sum(as.numeric(all_state_fund$`Sur / (Def)`), na.rm = TRUE)
  
  # Calculate Other EXP projections
  non_salary_proj_remaining_balance <- sum(as.numeric(all_state_fund$`Remaining Balance`), na.rm = TRUE)
  non_salary_proj_projected_value <- sum(as.numeric(all_state_fund$`Projected Y|E Value`), na.rm = TRUE)
  non_salary_proj_sur_def <- sum(as.numeric(all_state_fund$`Sur / (Def)`), na.rm = TRUE)
  
  # Store results for this cost center in a data frame
  alys_sf_summary <- data.frame(
    `Cost Center` = cost_center,
    `Remaining Balance - Salaries/OPC` = salary_proj_remaining_balance,
    `Projected YE Value - Salaries/OPC` = salary_proj_projected_value,
    `Sur/(Def) - Salaries/OPC` = salary_proj_sur_def,
    `Remaining Balance - Other EXP` = non_salary_proj_remaining_balance,
    `Projected YE Value - Other EXP` = non_salary_proj_projected_value,
    `Sur/(Def) - Other EXP` = non_salary_proj_sur_def
  )
  
  # Store this cost center summary in the results list
  alys_sf_final_tbl <- alys_sf_summary |> 
    summarise(
      `Salaries/OPC Projections` = sum(`Remaining.Balance...Salaries.OPC`, na.rm = TRUE) +
                                   sum(`Projected.YE.Value...Salaries.OPC`, na.rm = TRUE) +
                                   sum(`Sur..Def....Salaries.OPC`, na.rm = TRUE),
      `Other EXP Projections` = sum(`Remaining.Balance...Other.EXP`, na.rm = TRUE) +
                                sum(`Projected.YE.Value...Other.EXP`, na.rm = TRUE) +
                                sum(`Sur..Def....Other.EXP`, na.rm = TRUE),
      `Total Projections` = `Salaries/OPC Projections` + `Other EXP Projections`
    )
  
  # Append the final table to the results list
  alys_state_fund_results[[as.character(cost_center)]] <- alys_sf_final_tbl
}

# --- Combine all results into a final dataframe ---
alys_state_fund_summary_df <- bind_rows(alys_state_fund_results) |> 
  slice(1)
```


### Total Projections equals to row 268 
This is for the Analysis Tables. Combining the general and state funds for the tpe table. 

```{r antpe, warning=FALSE}
# --- bind the rows together ---

alys_tpe_summary <- bind_rows(alys_general_fund_summary_df, alys_state_fund_summary_df) 

# --- create the final TPE table ---

alys_tpe_final_tbl <- alys_tpe_summary |> 
  summarise(
    `Salaries/OPC Projections` = sum(as.numeric(`Salaries/OPC Projections`), na.rm = TRUE),
    `Other EXP Projections` = sum(as.numeric(`Other EXP Projections`), na.rm = TRUE),
    `Total Projections` = `Salaries/OPC Projections` + `Other EXP Projections`
  )
```



*Fourth calculation include the Percentage Table*

### Percentage Calculations
Here is a table that will calculate the percentages for the Percentage of Budget Depleted YTD Table.
Using the `pull` function which ensures a single numeric value is extracted rather than a dataframe column.

```{r percentages, warning=FALSE}
# finding the percentage of the gyltd_tpe and tpe_rb final tables

pcnt_final_tbl <- tibble(
  `Total Salaries` = round((pull(glytd_tpe_final_tbl, Salaries.OPC.Projections)) / 
                           (pull(tpe_rb_final_tbl, Salaries.OPC.Projections)) * 100, 0), # ensures that the output is rounded to the next whole num
  `Total Expenses` = round((pull(glytd_tpe_final_tbl, Other.EXP.Projections)) / 
                           (pull(tpe_rb_final_tbl, Other.EXP.Projections)) * 100, 0),
  `Overall Total`  = round((pull(glytd_tpe_final_tbl, Total.Projections)) / 
                           (pull(tpe_rb_final_tbl, Total.Projections)) * 100, 0)
)
```



#####
### Plotting the data for visulzations


The tables that have the name `final` in them will be used for the plots. These plots will serve as a way to make the data represent what the tables are showing within the main dataset. 

*First table compares the General Fund 1001 and the GL YTD EXP General Fund*
*Second table compares the Other Funds - 2089, 4000,  and 5000 and the GL YTD EXP State Grants*
*Third table compares the Total Projections Equals to Row 268 and GL YTD EXP General Fund 1001*
*Fourth table shows the Percent of Budget Depleted YTD Table Data*
*Fifth table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis table General Fund 1001*
*Sixth table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis table Other Funds - 2089, 4000,  and 5000 table*
*Seventh table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis Total Projections equals to row 268 table*
#####



### General Funds 1001 and Generl Funds 1001 GL YTD EXP comparison plot
The plot below will compare the GF and the GLYTD EXP General Funds 1001 data together. The `gf_glytd_comparison_data` includes the spending categories that are within the General Funds 1001. This will show which if the budget is close or over for the total overall budget of the spending categories. 

*First table compares the General Fund 1001 and the GL YTD EXP General Fund*

```{r gfirstplot}
# format in long 
gf_glytd_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), each = nrow(general_fund_summary_df) * 2),
  Fund = rep(c("1001 General Funds", "GL YTD 1001 General Funds"), each = nrow(general_fund_summary_df), times = 2),
  Value = c(general_fund_summary_df$Salaries.OPC.Projections, glytd_general_fund_summary_df$Salaries.OPC.Projections,
            general_fund_summary_df$Other.EXP.Projections, glytd_general_fund_summary_df$Other.EXP.Projections)
)

# Step 2: Sum the Revised Budget from both of the salary and non salary rows because these contain the revised budget totals
total_projections_gf <- sum(as.numeric(gf_salary_proj), na.rm = TRUE) + 
                        sum(as.numeric(gf_non_salary_proj), na.rm = TRUE)


# Step 3: Create the stacked horizontal bar chart with two threshold lines
ggplot(gf_glytd_comparison_data, aes(x = Category, y = Value, fill = Fund)) +
  geom_bar(stat = "identity", position = "stack") +  # Stacked bars
  geom_segment(aes(x = 1, xend = 2.5, y = total_projections_gf, yend = total_projections_gf), 
               linetype = "dashed", color = "blue", size = 1) +  # Short horizontal line for General Fund
  annotate("text", x = 2.5, y = total_projections_gf, 
           label = paste("Revised Budget", label_dollar()(total_projections_gf)), 
           color = "blue", vjust = -0.5, size = 6) +  # Label for General Fund line with formatted dollar value
  coord_flip() +  # Flip to make bars horizontal
  labs(title = "1001 General Funds and General Ledger YTD Comaparisons",
       subtitle = "All Cost Centers Under the 1001 General Funds",
       x = "Category", y = "Amount") +
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis labels as dollar values
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 20),  # Increase legend text size
        plot.title = element_text(hjust = 0.5, size = 18),  # Increase plot title size
        plot.subtitle = element_text(hjust = 0.5, size = 18),
        axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
        axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
        axis.text.x = element_text(size = 16),  # Increase x-axis labels size
        axis.text.y = element_text(size = 16))  # Increase y-axis labels size


# save the plot
ggsave("Charts/all_gf_glytd_comparison_data.png",width = 24, height = 16, dpi = 300)  
```


### 5000 State Grants Fund Other Funds - 2089, 4000, and 5000 and GL YTD EXP comparison plot
The plot below will compare the State Grants Fund and the GLYTD EXP 5000 State Grants data together. The `sg_glytd_comparison_data` includes the spending categories that are within the 5000 State Grants Fund. This will show which if the budget is close or over for the total overall budget of the spending categories. 

*Second table compares the Other Funds - 2089, 4000,  and 5000 and the GL YTD EXP State Grants*

```{r sf_sfglyt_plot, warning=FALSE}

# format in long 
sg_glytd_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), each = nrow(state_fund_summary_df) * 2),
  Fund = rep(c("5000 State Grants Funds", "GL YTD 5000 State Grants Funds"), each = nrow(state_fund_summary_df), times = 2),
  Value = c(
    state_fund_summary_df$Salaries.OPC.Projections, glytd_state_fund_summary_df$Salaries.OPC.Projections,
    state_fund_summary_df$Other.EXP.Projections, glytd_state_fund_summary_df$Other.EXP.Projections
  )
)


# Step 2: Sum the Revised Budget from both of the salary and non salary rows because these contain the revised budget totals
total_projections_sg <-  sum(as.numeric(sf_salary_proj), na.rm = TRUE) + 
                        sum(as.numeric(sf_non_salary_proj), na.rm = TRUE)


# Step 3: Create the stacked horizontal bar chart with two threshold lines
ggplot(sg_glytd_comparison_data, aes(x = Category, y = Value, fill = Fund)) +
  geom_bar(stat = "identity", position = "stack") +  # Stacked bars
  geom_segment(aes(x = 1, xend = 2.5, y = total_projections_sg, yend = total_projections_sg), 
               linetype = "dashed", color = "blue", size = 1) +  # Short horizontal line for General Fund
  annotate("text", x = 2.5, y = total_projections_sg, 
           label = paste("Revised Budget", label_dollar()(total_projections_sg)), 
           color = "blue", vjust = -0.5, size = 6) +  # Label for General Fund line with formatted dollar value
  coord_flip() +  # Flip to make bars horizontal
  labs(title = "Other Funds - 2089, 4000, and 5000 and General Ledger YTD Comparisons",
       subtitle = "5000 State Grant Funds",
       x = "Category", y = "Amount") +
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis labels as dollar values
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 20),  # Increase legend text size
        plot.title = element_text(hjust = 0.5, size = 18),  # Increase plot title size
        plot.subtitle = element_text(hjust = 0.5, size = 18),
        axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
        axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
        axis.text.x = element_text(size = 16),  # Increase x-axis labels size
        axis.text.y = element_text(size = 16))  # Increase y-axis labels size

# save the plot
ggsave("Charts/all_sg_glytd_comparison_data.png",width = 24, height = 18, dpi = 300)  
```


### Total Projections Equals to Row 268 and GL YTD EXP - 1001 General Funds and 5000 State Grants comparison plot
The plot below will compare the General Grants Fund and the GLYTD EXP 5000 State Grants data together. The `tpe_glytd_comparison_data` includes the spending categories that are within the 1001 General Grants and 5000 State Grants Fund. This will show which if the budget is close or over for the total overall budget of the spending categories. 

*Third table compares the Total Projections Equals to Row 268 and GL YTD EXP General Fund 1001*


```{r tpe_plot, warning=FALSE}
# Step 1: Construct Comparison Data Frame Correctly
tpe_glytd_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), times = 2),  # Repeat twice per fund
  
  Fund = rep(c("1001 General Grants Fund", "GL YTD EXP 1001 General Grants Fund",
               "5000 State Grants Fund", "GL YTD EXP 5000 State Grants Fund"), each = 2),  # Repeat each twice
  
  Value = c(
    tpe_rb_final_tbl$Salaries.OPC.Projections, glytd_tpe_final_tbl$Salaries.OPC.Projections,
    tpe_rb_final_tbl$Other.EXP.Projections, glytd_tpe_final_tbl$Other.EXP.Projections
  )  # Ensure values align correctly
)

# Step 2: Sum the Revised Budget Correctly
total_projections_tpe <- sum(as.numeric(tpe_rb_final_tbl$Total.Projections), na.rm = TRUE) + 
                         sum(as.numeric(glytd_tpe_final_tbl$Total.Projections), na.rm = TRUE)


# Step 3: Create the stacked horizontal bar chart with two threshold lines
ggplot(tpe_glytd_comparison_data, aes(x = Category, y = Value, fill = Fund)) +
  geom_bar(stat = "identity", position = "stack") +  # Stacked bars
  geom_segment(aes(x = 1, xend = 2.5, y = total_projections_tpe, yend = total_projections_tpe), 
               linetype = "dashed", color = "blue", size = 1) +  # Short horizontal line for General Fund
  annotate("text", x = 2.5, y = total_projections_tpe, 
           label = paste("Revised Budget", label_dollar()(total_projections_tpe)), 
           color = "blue", vjust = -0.5, size = 6) +  # Label for General Fund line with formatted dollar value
  coord_flip() +  # Flip to make bars horizontal
  labs(title = "Total Projections Equals to Row 268 and GL YTD EXP Comaparisons",
       subtitle = "1001 General Funds and 5000 State Grant Funds",
       x = "Category", y = "Amount") +
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis labels as dollar values
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 20),  # Increase legend text size
        plot.title = element_text(hjust = 0.5, size = 18),  # Increase plot title size
        plot.subtitle = element_text(hjust = 0.5, size = 18),
        axis.title.x = element_text(size = 16, face = "bold"),  # Increase x-axis title size
        axis.title.y = element_text(size = 16, face = "bold"),  # Increase y-axis title size
        axis.text.x = element_text(size = 16),  # Increase x-axis labels size
        axis.text.y = element_text(size = 16))  # Increase y-axis labels size

# save the plot
ggsave("Charts/all_tpe_glytd_comparison_data.png",width = 24, height = 16, dpi = 300)  
```


### Percent of Budget Depleted YTD Table Data plot
The plot below will compare the glytd and tp data in the Percentage of Budget Depleted YTD comparison. The `pcnt_final_data` includes the spending categories that are within the 1001 General Grants Fund. This will show which if the budget is close or over for the total overall budget of the spending categories. 

*Fourth table shows the Percent of Budget Depleted YTD Table Data*


```{r pct_plot, warning=FALSE}
# create a percentage comparison dataframe
all_pcnt_final_data <- data.frame(
  Category = rep(c("Total Salaries", "Total Expenses", "Overall Total"), each = nrow(pcnt_final_tbl)),  
  Fund = rep("GL YTD EXP 1001 General & 5000 State Grants Funds", nrow(pcnt_final_tbl) * 3),  # Repeat the fund name for each category
  Value = round(c(
    (glytd_tpe_final_tbl$Salaries.OPC.Projections / tpe_rb_final_tbl$Salaries.OPC.Projections) * 100,  
    (glytd_tpe_final_tbl$Other.EXP.Projections / tpe_rb_final_tbl$Other.EXP.Projections) * 100,
    (glytd_tpe_final_tbl$Total.Projections / tpe_rb_final_tbl$Total.Projections) * 100
  ), 0)  # Round to whole numbers
)


# Step 2: Calculate the total revised budget for the first rows defined in the dataset
# Sum the Revised Budget from both gf_first_rows and gf_remaining_rows
total_projections_sg <- sum(as.numeric(sf_salary_proj), na.rm = TRUE) + 
                        sum(as.numeric(sf_non_salary_proj), na.rm = TRUE)


# Step 3: Create the stacked horizontal bar chart with two threshold lines
ggplot(all_pcnt_final_data, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +  # Side-by-side bars (non-stacked)
  labs(title = "Percentage of Budget Depleted YTD Comparisons",
       subtitle = "1001 General Funds and 5000 State Grant Funds",
       x = "Category", y = "Percentage (%)") +
  coord_flip()+
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Format y-axis labels as percentages
  scale_fill_manual(values = c("Total Salaries" = "#004D40", 
                               "Total Expenses" = "#1E88E5", 
                               "Overall Total" = "#5D3A9B")) +  # Custom colors for each category
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 14),  # Increase legend text size
        plot.title = element_text(hjust = 0.5, size = 16),  # Increase plot title size
        plot.subtitle = element_text(hjust = 0.5,size = 12),
        axis.title.x = element_text(size = 14, face = "bold" ),  # Increase x-axis title size
        axis.title.y = element_text(size = 14, face = "bold"),  # Increase y-axis title size
        axis.text.x = element_text(size = 14),  # Increase x-axis labels size
        axis.text.y = element_text(size = 14))  # Increase y-axis labels size


# save the plot
ggsave("Charts/all_pct_tbl.png",width = 18, height = 16, dpi = 300)  
```


*Fifth table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis table General Fund 1001*

### Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis table General Fund 1001 comparison plot
The plot below will compare the Analysis table General Fund data together. The `ays_gf_comparison_data` includes the spending categories that are within the 1001 General Grants. This will show which if the budget is close or over for the total overall budget of the spending categories. 


```{r anlys_gf_plot, warning=FALSE}
# Create the comparison data for plotting
ays_gf_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), each = nrow(alys_general_fund_summary_df)),  
  Fund = rep("1001 General Grants Fund", times = 2 * nrow(alys_general_fund_summary_df)),  
  Value = c(
    alys_general_fund_summary_df$`Salaries/OPC Projections`,
    alys_general_fund_summary_df$`Other EXP Projections`
  )
)


# Step 2: Create the non-stacked horizontal bar chart with custom colors and bold axis titles
ggplot(ays_gf_comparison_data, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +  # Non-stacked bars (side-by-side)
  scale_fill_manual(values = c("Salaries/OPC Projections" = "steelblue", 
                               "Other EXP Projections" = "darkgreen")) +  # Custom colors
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis values with dollar signs and commas
  coord_flip()+
  labs(title = "Analysis Table 1001 General Grant Funds Comparisons",
       subtitle = "1001 General Funds",
       x = "Category",
       y = "Amount") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 16),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.title.x = element_text(size = 18, face = "bold"),  # Bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Bold y-axis title
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 18)
  )

# save the plot
ggsave("Charts/all_ays_gf_comparison_data.png",width = 18, height = 16, dpi = 300)  
```


 
*Sixth table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis Other Funds - 2089, 4000, and 5000 table*

### Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis Other Funds - 2089, 4000,  and 5000 table plot
The plot below will compare the Analysis table General Fund data together. The `ays_sg_comparison_data` includes the spending categories that are within the 5000 State Grants Fund. This will show which if the budget is close or over for the total overall budget of the spending categories. 


```{r anlys_gf_plot, warning=FALSE}
# Create the comparison data for plotting
ays_sf_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), each = nrow(alys_state_fund_summary_df)),  
  Fund = rep("1001 General Grants Fund", times = 2 * nrow(alys_state_fund_summary_df)),  
  Value = c(
    alys_state_fund_summary_df$`Salaries/OPC Projections`,
    alys_state_fund_summary_df$`Other EXP Projections`
  )
)

# Step 2: Create the non-stacked horizontal bar chart with custom colors and bold axis titles
ggplot(ays_sf_comparison_data, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +  # Non-stacked bars (side-by-side)
  scale_fill_manual(values = c("Salaries/OPC Projections" = "steelblue", 
                               "Other EXP Projections" = "darkgreen")) +  # Custom colors
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis values with dollar signs and commas
  coord_flip()+
  labs(title = "Analysis Table Other Funds - 2089, 4000, and 5000 Comparisons",
        subtitle = "5000 State Grant Funds",
       x = "Category",
       y = "Amount") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 16),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.title.x = element_text(size = 18, face = "bold"),  # Bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Bold y-axis title
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 18)
  )
# save the plot
ggsave("Charts/all_ays_sg_comparison_data.png",width = 18, height = 16, dpi = 300)  
```



*Seventh table compares the Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis Total Projections equals to row 268 table*

### Remaining Balance, Sur/ Def, and Projected Y|E Value for the Analysis Total Projections equals to row 268 plot
The plot below will compare the Analysis table General Fund data together. The `aystpe_comparison_data` includes the spending categories that are within the 1001 General Grants. This will show which if the budget is close or over for the total overall budget of the spending categories. 


```{r anlys_tpe_plot, warning=FALSE}
# Create the comparison data for plotting
aystpe_comparison_data <- data.frame(
  Category = rep(c("Salaries/OPC Projections", "Other EXP Projections"), times = 2),  # Repeat twice per fund
  
  Fund = rep(c("1001 General Grants Fund", "GL YTD EXP 1001 General Grants Fund",
               "5000 State Grants Fund", "GL YTD EXP 5000 State Grants Fund"), each = 2),  # Repeat each twice
  
  Value = c(
    ays_gf_comparison_data$Category, glytd_tpe_final_tbl$Salaries.OPC.Projections,
    ays_gf_comparison_data$Other.EXP.Projections, glytd_tpe_final_tbl$Other.EXP.Projections
  )  # Ensure values align correctly
)

# Step 2: Create the non-stacked horizontal bar chart with custom colors and bold axis titles
ggplot(aystpe_comparison_data, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +  # Non-stacked bars (side-by-side)
  scale_fill_manual(values = c("Salaries/OPC Projections" = "steelblue", 
                               "Other EXP Projections" = "darkgreen")) +  # Custom colors
  scale_y_continuous(labels = label_dollar()) +  # Format y-axis values with dollar signs and commas
  coord_flip()+
  labs(title = "Analysis Table Total Projections Equals to Row 268 Comparisons",
       subtitle = "1001 General Funds and 5000 State Grant Funds",
       x = "Category",
       y = "Amount") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 16),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.title.x = element_text(size = 18, face = "bold"),  # Bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Bold y-axis title
    axis.text.x = element_text(size = 18, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 18)
  )

# save the plot
ggsave("Charts/all_ays_tpe_comparison_data.png",width = 18, height = 16, dpi = 300)  
```































































































































