---
title: "Gl_and_Budget_vs_Actuals"
author: "Selenea Gibson"
date: "2025-01-02"
output: html_document
---

## Project Overview
This markdown file will use the DHCD Finance dept documents that are for reporting expenditures and GL data. The data will 

```{r libraries}
# libraries
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(DT)
```

## Data 
Reading in the files for the data that will be used. Put new data files in the "Data" folder and replace the file names below with the names of the new files.

```{r data}

## change the names of these files

GL_file_name <- "GL-Operating Actual Expenses YTD 2024-12-19 13_03 EST.xlsx"

budg_vs_act_file_name <- "Budget_vs_Actuals_-_BBMR YTD 2024-12-19.xlsx"


## 
## read data from files


gl_data <- readxl::read_xlsx(here("Data", GL_file_name), 
                             skip = 13,
                             col_types = c(rep("text",2),   
                                           "date", 
                                           rep("text",6),
                                           rep("numeric", 3),
                                           rep("text",26)))


## the first row of the data file is mostly empty except for a few important
##dates. Read those data and do some formatting

first_row_of_data <- readxl::read_xlsx(here("Data", budg_vs_act_file_name)) |> 
  names() 


name_prefixes <- tibble(first_row_of_data) |> 
  mutate(prefixes = ifelse(str_detect(file_headers, "\\.\\."), 
                          NA_character_, 
                          file_headers)) |> 
fill(prefixes, .direction = "down") |> 
  pull(prefixes)


### read the data file again but skip the first line

budg_vs_actuals <- readxl::read_xlsx(here("Data", budget_vs_actuals_file_name),
                  skip = 1)


### the names in the file repeat themselves, but this combines the dates 
### with the repetative names

for(i in seq_along(name_prefixes)){
  plain_name <- gsub("\\.\\..*","", names(budg_vs_actuals)[i]) 
  
  names(budg_vs_actuals)[i] <- 
    ifelse(is.na(name_prefixes[i]) & str_length(plain_name) > 0,
           plain_name,
           paste(plain_name,name_prefixes[i], sep = "_"))
}

 names(budg_vs_actuals)

```

## Cleaning the datasets
I need to remove some of the rows that are not needed for the analysis. I will then make the new row be the header row. 

*NOTE THERE IS A COLUMN NAMED NA- REMOVE THIS AS THERE IS NO DATA IN THIS COLUMN?* 

```{r cleaningdata}


## add round all expenditure days to the expenditure month
gl_data <- gl_data |> 
  mutate(expend_month = as.Date(floor_date(`Accounting Date` , unit = "month")))


### remove columns that have nulls for all values, replace "blank" with an actual 'null' value

budg_vs_actuals <- budg_vs_actuals |> 
  select_if(~sum(!is.na(.)) > 0) |> 
  mutate(Grant = ifelse(Grant == "(Blank)", NA_character_, Grant))

### create a dataframe of overall expenditures by worktags from the 
### budget vs actuals file

budg_vs_actuals_overall <- 
  budg_vs_actuals |> 
  select(-matches("_")) 

### create a dataframe of monthly expenditures by worktags from the 
### budget vs actuals file. Change the table from wide to long.

budget_vs_Actuals_monthly <- 
  budg_vs_actuals |> 
  select(-matches("Budget|total",ignore.case = T)) |> 
  pivot_longer(cols = -!matches("_"), names_to = "date_topic", values_to = "amounts") |> 
  separate(date_topic, sep = "_", into = c("acct_category", "month")) |> mutate(month = as.Date(paste0(month,"-01")))


```



### monthly amounts match

check to see if the monthly expenditures for each categories match between the two files.


```{r}

### aggregate the ledger data by month and worktags
monthly_spendies <- gl_data |> 
  group_by(expend_month, Fund, `Cost Center`, Grant, 
           `Spend Category as Worktag`) |> 
  reframe(total_spend_on_gl = sum(Amount, na.rm = T))

### compare between two files
monthly_spendies_compared <- budget_vs_Actuals_monthly |> 
  filter(acct_category == "Actuals") |> 
  full_join(monthly_spendies, by = c("Cost Center"="Cost Center",
                                     "Fund"="Fund",
                                     "Grant" = "Grant",
                                     "Spend Category" = "Spend Category as Worktag",
                                     "month"="expend_month")) 

### filter on those where values dont match
mismatched_monthly_expenditures <- monthly_spendies_compared |> 
  filter(round(amounts) != round(total_spend_on_gl) ) 


message(scales::percent(nrow(mismatched_monthly_expenditures)/nrow(monthly_spendies_compared), accuracy = .1), " of rows are mismatched")


```

### overall expenditures

check to see if the total expenditures for each category match between the two files.

```{r}

overall_spendies <- gl_data |> 
  group_by(Fund, `Cost Center`, Grant, `Spend Category as Worktag`) |> 
  reframe(total_spend_on_gl = sum(Amount, na.rm = T),
          number_of_lines = n())


overall_spendies_compared <- budg_vs_actuals_overall |> 
    full_join(overall_spendies, by = c("Cost Center"="Cost Center",
                                     "Fund"="Fund",
                                     "Grant" = "Grant",
                                     "Spend Category" = "Spend Category as Worktag")) 


mismatched_overall_expenditures <- overall_spendies_compared |> 
  filter(round(`Total Spent` ) != round(total_spend_on_gl) ) |> 
  mutate_at(vars(matches("total", ignore.case = T)), scales::dollar)


message(scales::percent(nrow(mismatched_overall_expenditures)/nrow(overall_spendies), accuracy = .1), " are mismatched")


```


## Join the reports 
I want to join the reports together based on a unique ID. I will have to create my own unique ID because the column names and row numbers do not match. 
Assigning the number to start at 1 for the id count. This will now allow for a proper join. 
*NOTE I HAVE .1 AT THE END OF SOME OF THE DUPLICATE COLUMNS. WANT TO REMOVE AND HAVE TRIED BUT NOTHING WORKED TO DO SO.*

```{r joins}




```


## Benchmarks
Make a color coded marking for spends that are over budget or near the budget spending mark. There will be numerous benchmarks. This will occur at 80% spent amount. 

Examples below

```{r benchmarks}
# make it numeric
merged_data$`Total Spent` <- as.numeric(merged_data$`Total Spent`)

# create the new column with 1 if the amount is greater than 80% of the total spend, 0 otherwise
merged_data$over_80_percent <- ifelse(merged_data$Amount > 0.8 * merged_data$`Total Spent`, 1, 0)

# create a new column that flags rows where total spend is negative
merged_data$total_spend_negative <- ifelse(merged_data$'Total Spent' < 0, 1,0)
```







## Mapping and plots
Making a map and a plot for the various data analysis. This can be altered to your liking and can include more data 

```{r plot}
# create a color column for plotting (orange if over 80%, default color otherwise)
merged_data$color <- ifelse(merged_data$over_80_percent == 1, "orange", "black")

# plot the data with color
ggplot(merged_data, aes(x = ID, y = Amount, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_identity() +
  theme_minimal() +
  labs(title = "Spending by ID",
       x = "ID", 
       y = "Amount")
```








































































































## Example code below

# the code below will read a Payroll Expense Ledger report
# from workday and create a summary excel file and a couple charts



# these will load a couple additional R tools for the work
# libraries are extra R tools
library(tidyverse)
library(scales)



# this will read the excel file. The file needs to be in the same
# folder as the R script. And the name in parenthesis needs to match
# the file name of the excel file
expenditure_data <-
  readxl::read_xlsx("June Payroll Expense Ledger by Org Excel 2024-07-17 10_21 EDT.xlsx",
                    skip = 8)




# This part aggregates the transaction in the GL file by service
# then sums the expenditures.

service_exp <- expenditure_data |>
  group_by(`Service (Cost Center Hierarchy)`) |>
  summarise(total_expenditures = sum(`Ledger Amount ($)`)) |>
  ungroup() |>
  arrange(desc(total_expenditures))

# you just created an object called "service_exp" which
# can be seen be executing this
print(service_exp)






# This part aggregates the transaction in the GL file by cost center
# then sums the expenditures.

cost_center_exp <- expenditure_data |>
  group_by(`Cost Center`) |>
  summarise(total_expenditures = sum(`Ledger Amount ($)`)) |>
  ungroup() |>
  arrange(desc(total_expenditures))

# you just created an object called "cost_center_exp" which
# can be seen be executing this
cost_center_exp


# this part creates an excel file with the costcenter
# and service expenditures

writexl::write_xlsx(list(`expend by service` = service_exp,
                         `expend by cost center` = cost_center_exp), "summarized_expenditures.xlsx")


#####
#####
### these parts make charts
#####
#####


# this part aggregates by date posted, summing
# expenditures byt the day the posted to the GL

daily_spend <- expenditure_data |>
  mutate(day = floor_date(as.Date(`Accounting Date` ), "day")) |>
  group_by(day) |>
  summarise(total_expenditures = sum(`Ledger Amount ($)`)) |>
  ungroup()

print(daily_spend)


### this part makes the chart

daily_spend |>
  ggplot(aes(day, total_expenditures))+
  geom_col() +
  theme_classic() +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  geom_hline(yintercept=0) +
  labs(title = "DGS Posted Expenditures throughout the month of June 2024",
       subtitle = "Expenditures by day",
       y = "Expenditures ($)",
       x = "Date posted on GL")

### this part saves the chart
ggsave("daily_spend.pdf",
       width = 8, height = 5)



## this part aggregates bby fund, summing
## expenditures, then makes a chart

expenditure_data |>
  separate(Fund, into = c("number", "Fund"), extra = "merge", remove = T) |>
  group_by(Fund) |>
  summarise(total_spend = sum(`Ledger Amount ($)`)) |>
  ggplot(aes(reorder(Fund, (total_spend)), total_spend)) +
  geom_col(fill = "orange") +
  theme_classic() +
  coord_flip() +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6), expand = c(0, 0), limits = c(0,3000000))+
  labs(title = "DGS Posted Expenditures by Fund",
       subtitle = "Posted in June 2024",
       y = "Expenditures ($)",
       x = "")


## this saves the chart as a pdf

ggsave("spend_by_fund.pdf",
       width = 8.5, height = 5)




## this part filters transaction to the manual JEs
### then aggregates by by fund summing the GL transactions
## by the fund

manual_change_data <- expenditure_data |>
  filter(str_detect(`Journal Source`, "Manual")) |>
  separate(Fund, into = c("number", "Fund"), extra = "merge", remove = T) |>
  group_by(Fund) |>
  summarise(total_spend = sum(`Ledger Amount ($)`))




### this part makes the chart showing where the manual
### GL's occur

ggplot() +
  geom_col(data = manual_change_data,
           mapping = aes(reorder(`Fund`, (total_spend)), total_spend),
           fill = "orange") +
  geom_text(data = manual_change_data,
            mapping = aes(reorder(`Fund`, (total_spend)), 0,
                          label = scales::dollar(total_spend/1000000, suffix = "M")), hjust = -.2) +
  theme_classic() +
  geom_hline(yintercept=0) +
  coord_flip() +
  geom_text() +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6), expand = c(0, 0))+
  labs(title = "Manual Accounting Adjustments",
       subtitle = "Made in June 2024",
       y = "Expenditures ($)",
       x = "")


## this saves the chart as a pdf
ggsave("manual_adjustments_by_fund.pdf",
       width = 8.5, height = 5)

