---
title: "Summer_food_program_Gl_and_Budget_vs_Actuals"
author: "Selenea Gibson & Jason Howard"
date: "2025-01-02"
output: html_document
---

# Project Overview
This markdown file will use the DHCD Finance dept documents that are for reporting expenditures and GL data. The data is collected from Workday.

```{r libraries, warning=FALSE, message=TRUE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
```


## Reading in the data 
Reading in the files for the data that will be used. Put new data files in the "Data" folder and replace the file names below with the names of the new files.

```{r data}
# get the latest files in the data folder
get_latest_file <- function(pattern) {
  files <- list.files(path = "Data", pattern = pattern, full.names = TRUE)
  return(files[order(file.info(files)$mtime, decreasing = TRUE)][1])
}

# assigning the names of the latest files to the new variables
gl_file_name <- basename(get_latest_file("GL-Operating Actual Expenses YTD .* EST\\.xlsx$"))
budg_vs_act_file_name <- basename(get_latest_file("Budget_vs_Actuals_-_BBMR YTD .*\\.xlsx$"))

# read data from files
gl_data <- readxl::read_xlsx(here("Data", gl_file_name), 
                             skip = 13,
                             col_types = c(rep("text",2),   
                                           "date", 
                                           rep("text",6),
                                           rep("numeric", 3),
                                           rep("text",26))) |> 
  mutate(`Accounting Date` = as.Date(`Accounting Date`))

# the first row of the data file is mostly empty except for a few important dates. 
# read those data and do some formatting to clean up the data
first_row_of_data <- 
  readxl::read_xlsx(here("Data", budg_vs_act_file_name)) |> 
  names()

# take out the data that is before the numeric values
name_prefixes <- 
  tibble(first_row_of_data) |> 
  mutate(prefixes = ifelse(str_detect(first_row_of_data, "\\.\\."), 
                           NA_character_, 
                           first_row_of_data)) |> 
  fill(prefixes, .direction = "down") |> 
  mutate(prefixes = ifelse(is.na(prefixes), 
                           NA_character_, 
                           paste0(prefixes, "-01"))) |> 
  pull(prefixes)

# read the data file again but skip the first line
budg_vs_actuals <- 
  readxl::read_xlsx(here("Data", budg_vs_act_file_name),
                    skip = 1)

# remove the total budget row as this is not needed 
# and will mess up the data. 
budg_vs_actuals <- budg_vs_actuals |> 
  filter(`Cost Center` != "Total") 

# the names in the file repeat themselves, but this combines the dates with the repetitive names
for(i in seq_along(name_prefixes)){
  plain_name <- gsub("\\.\\..*","", names(budg_vs_actuals)[i]) 
  
  names(budg_vs_actuals)[i] <- 
    ifelse(is.na(name_prefixes[i]) & str_length(plain_name) > 0,
           plain_name,
           paste(plain_name,name_prefixes[i], sep = "_"))
}

# remove commitments and obligations since we dont need them
budg_vs_actuals <- budg_vs_actuals |> 
  select(-matches("^Commitments|^Obligations"))

# looking at the new names 
names(budg_vs_actuals)
```


## Cleaning the datasets
I need to remove some of the rows that are not needed for the analysis. I will then make the new row be the header row. 

```{r cleaningdata}
# add round all expenditure days to the expenditure month
gl_data <- gl_data |> 
  mutate(expend_month = floor_date(`Accounting Date` , unit = "month"))

# remove columns that have nulls for all values 
# replace "blank" with an actual 'null' value
budg_vs_actuals <- budg_vs_actuals |> 
  select_if(~sum(!is.na(.)) > 0) |> 
  mutate(Grant = ifelse(Grant == "(Blank)", NA_character_, Grant))

```


####
## Creating the fiscal months
This will create the fiscal months and year that can be used for `754 SFSP Expense Report`. 

```{r, warning=FALSE}
# -- identifying the months of the fiscal year ---
# find all months of fiscal year
all_column_names <- names(budg_vs_actuals)
month_columns <- all_column_names[str_detect(all_column_names, "Actuals_")]
get_first_date <- as.Date(paste0(strsplit(month_columns, "_")[[1]][2],"-01"))
all_year_months <- seq(get_first_date, get_first_date +months(11), by = "month")

# --- placeholders for the missing months ---
# add missing months into the table
for (col in paste0("Actuals_", all_year_months)) {
  if (!col %in% names(budg_vs_actuals)) {
    budg_vs_actuals[[col]] <- NA_real_  # Add missing columns filled with NA
  }
}

# Build the expense report table dynamically
expense_report_tbl <- budg_vs_actuals |> 
  group_by(Fund, `Spend Category`, `Cost Center`) |> 
  summarise(
    across(where(is.numeric), ~ sum(.x, na.rm = TRUE)),  # Sum Actuals columns
    .groups = "drop"
  )


# Rename columns to the "Month Year" format (e.g., "June 2024")
expense_report_tbl <- expense_report_tbl |> 
  rename_with(
  ~ as.character(month(as.Date(gsub("Actuals_","", .x)), label = T, abbr = F)),
  starts_with("Actuals_")
)
```


###
## Calculations for workbook tables
Calculating the gl_ytd_exp column. I will also rename the `Total Actual` column to the `Total Expenditure` column. Then join the gl table to the expense report table. Then calculate the columns that are needed for the analysis table. 

```{r calculations}
# --- aggregate the GL_YTD_EXP by Fund, Spend Category, and Cost Center ---
gl_data_exp <- gl_data |> 
  group_by(Fund, `Spend Category as Worktag`, `Cost Center`) |> 
  summarise(`GL YTD EXP` = sum(Amount, na.rm = TRUE), .groups = "drop")

# --- Joining datasets ---
# joining the GL YTD data to the expense_report_tbl
expense_report_tbl <- expense_report_tbl |> 
  left_join(gl_data_exp, 
            by = c("Spend Category" = "Spend Category as Worktag", 
                   "Cost Center" = "Cost Center", 
                   "Fund" = "Fund")) |> 
  rename(`Total Expenditures` = `Total Actuals`) |> 
  mutate(`GL YTD EXP` = if_else(is.na(`GL YTD EXP`), 0, `GL YTD EXP`)) 
```


1. Create the summary table that houses the summer food expense data.

```{r filteringsumprogram}
# --- create the all_expense_rep dataframe  ---

# creating a summary table for the final output
all_expense_rep <- expense_report_tbl |> 
  arrange(Fund, `Cost Center`, `Spend Category`) |> 
  select(-c(
    `Adopted Budget`,
    `Corrections to Budget`,
    Amendments,
    `Total Spent`
  ))

# filter to show the summer_food_program only have to put the full name into the filter
summer_expense_rep <- all_expense_rep |> 
  filter(str_detect(`Cost Center`, "Summer")) |> 
  relocate(`Spend Category`, .after = `Cost Center`) |> 
  relocate(`Total Expenditures`, .after = `GL YTD EXP`) |> 
  mutate(`GL YTD EXP` = round(`GL YTD EXP`, 2))  # rounds to 2 decimal places
```


2. Calculating the columns for the Analysis Table
Create the columns that are needed for the second table that is in the `754 SFSP Expense Report`.

```{r calculatingcols, warning=FALSE}
# create a new dataframe with only the new columns included and not the rest
# tblwhodis is a placeholder for the new calculated columns

# --- calculate the Remaining Balance column ---

# create a new column for the remaining balance and make the gl ytd exp numeric
tblwhodis <- summer_expense_rep |> 
  mutate(
    `Revised Budget` = as.numeric(gsub("[^0-9.-]", "", `Revised Budget`)),
    `GL YTD EXP` = as.numeric(gsub("[^0-9.-]", "", `GL YTD EXP`)),
    `Remaining Balance` = ifelse(
      is.na(`Revised Budget`) | is.na(`GL YTD EXP`), 
      NA,  
      `Revised Budget` - `GL YTD EXP`  
    )
  )

# --- create a numerical fiscal month ---

latest_fiscal_month <- month(max(all_year_months, na.rm = TRUE)) 

# --- Projected Y|E Value calculation ---

# calculating the projected Y|E Values
tblwhodis <- tblwhodis |> 
  mutate(
    FiscalMonth = latest_fiscal_month,  
    `Projected Y|E Value` = ifelse(
      is.na(`GL YTD EXP`) | is.na(FiscalMonth), 
      NA,  
      (`GL YTD EXP` / FiscalMonth) * 12  
    )
  ) |> 
  select(-FiscalMonth)  # Drop FiscalMonth column while keeping everything else

# --- Sur / (Def) column ---

# calculate the Sur / (Def) column 
tblwhodis <- tblwhodis |>
  mutate(
    `Sur / (Def)` = ifelse(
      is.na(`Revised Budget`) | is.na(`Projected Y|E Value`),
      NA,  
      `Revised Budget` - `Projected Y|E Value`  
    )
  )


# --- Notes column ---# ---`Projected Y|E Value` Notes column ---

# here we will make a notes column that can be used in excel
# initialize as empty character column
tblwhodis$Notes <- NA_character_  
```


####
### Creating the Excel workbooks
####

## Create the Excel workbooks for wb1 and wb2
Creating wb1 for the summer food program and wb2 for the analysis table.

```{r wbs}
summer_expense_rep <- summer_expense_rep |> 
  group_by(Fund, `Cost Center`) |> 
  mutate(row_num = row_number()) |> 
  ungroup() |> 
  mutate(
    Fund = ifelse(row_num == 1, Fund, ""),
    `Cost Center` = ifelse(row_num == 1, `Cost Center`, "")
  ) |> 
  select(-row_num)

# only keep the columns that are needed for this table
analysis_tbl <- tblwhodis |> 
  select(`Remaining Balance`, `Projected Y|E Value`, `Sur / (Def)`, Notes)
```


###
## Styling for the Excel workbooks
This will create the styling for the 2 wbs. The Summer Food Program and the Analysis Table.
*NOTE: You do NOT need to change the name of the title header for the tables. As the function in line 294 will automatically detect the fiscal year end or beginning date and apply the month/day to the header (line 304)*

```{r stylingforwb1}
# --- get the current system date when this was run ---

current_date <- Sys.Date()

# extract year and month
current_year <- format(current_date, "%Y")
current_month <- as.numeric(format(current_date, "%m"))


# Determine the fiscal year
if (current_month >= 7) {
  fiscal_year <- as.numeric(current_year) + 1
} else {
  fiscal_year <- as.numeric(current_year)
}

# --- create a workbook ---

# format the title header 

title_header <- paste0("FY ", fiscal_year, " Budget and Actual as of ", format(current_date, "%m/%d/%Y"))

# add the worksheet to the created workbook
wb1 <- createWorkbook()
addWorksheet(wb1, "Summer Expense Report")

# --- add Title Header (Row 1) ---

# creating a title header 
writeData(wb1, "Summer Expense Report", title_header, startRow = 1, startCol = 1)
header_style <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#ED7D31")

# merge the cells for the title
mergeCells(wb1, "Summer Expense Report", cols = 1:ncol(summer_expense_rep), rows = 1)
addStyle(wb1, "Summer Expense Report", header_style, rows = 1, cols = 1:ncol(summer_expense_rep), gridExpand = TRUE)

# --- write the Data (Row 2) with Column Names ---

writeData(wb1, "Summer Expense Report", summer_expense_rep, startRow = 2, startCol = 1, colNames = TRUE)

# --- align for Column Headers (Row 2) ---

center_style <- createStyle(halign = "center", valign = "center")
addStyle(wb1, "Summer Expense Report", center_style, rows = 2, cols = 1:ncol(summer_expense_rep), gridExpand = TRUE)

# --- style the Column Names (Row 2) ---

column_names_style <- createStyle(
  halign = "center", 
  valign = "center", 
  textDecoration = "bold", 
  fgFill = "#ED7D31" 
)

# apply the style to the column names (row 2)
addStyle(wb1, "Summer Expense Report", column_names_style, rows = 2, cols = 1:ncol(summer_expense_rep), gridExpand = TRUE)

# --- saving the wb ---
saveWorkbook(wb1, "summer_expense_report.xlsx", overwrite = TRUE)
```


Styling for the 2 wb or Analysis table.

```{r stylingforwb2}
# only keep the columns that are needed for this table
analysis_tbl <- tblwhodis |> 
  select(`Remaining Balance`, `Projected Y|E Value`, `Sur / (Def)`, Notes)

# --- create a workbook ---

# add the worksheet to the created workbook
wb2 <- createWorkbook()
addWorksheet(wb2, "analysis_tbl")

# --- write data to the worksheet 2 ---

writeData(wb2, "analysis_tbl", analysis_tbl, startRow = 2, startCol = 1, colNames = TRUE)

# --- header for the table---

# add a header, style, and color with the label "Analysis" in the first row
writeData(wb2, "analysis_tbl", "ANALYSIS", startRow = 1, startCol = 1)
header_style <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4")

# merge the cells
mergeCells(wb2, "analysis_tbl", cols = 1:ncol(analysis_tbl), rows = 1)
addStyle(wb2, "analysis_tbl", header_style, rows = 1, cols = 1:ncol(analysis_tbl), gridExpand = TRUE)

# --- align for column headers ---

# center alignment for column headers
center_style <- createStyle(halign = "center", valign = "center")
addStyle(wb2, "analysis_tbl", center_style, rows = 2, cols = 1:ncol(analysis_tbl), gridExpand = TRUE)

# --- style the column names (e.g., "Remaining Balance", "Sur / (Def)", etc.) ---

# styling the column names for the datatable
column_names_style <- createStyle(
  halign = "center", 
  valign = "center", 
  textDecoration = "bold", 
  fgFill = "#C6E0B4" 
)

# apply the style to the column names (first row of data)
addStyle(wb2, "analysis_tbl", column_names_style, rows = 2, cols = 1:ncol(analysis_tbl), gridExpand = TRUE)

# --- saving the wb

saveWorkbook(wb2, "analysis_tbl.xlsx", overwrite = TRUE)
```


###
## Combining the workbooks
Last step, "Whew!", is to join the two of these workbooks together to have 1 full working workbook that can be used for the team. 

```{r combiningwbs}
# --- combine the two workbooks on the same worksheet with full formatting ---

# create a new workbook
wb_combined <- createWorkbook()

# add a worksheet for the combined data
addWorksheet(wb_combined, "tables_unite")

# define styles for titles and headers
title_style_summer <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#FFA500")
title_style_analysis <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4")
header_style_summer <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#FFA500")
header_style_analysis <- createStyle(halign = "center", valign = "center", textDecoration = "bold", fgFill = "#C6E0B4")

# write the Summer Expense Report table with title and formatting
writeData(wb_combined, "tables_unite", "Summer Expense Report", startRow = 1, startCol = 1)
mergeCells(wb_combined, "tables_unite", cols = 1:ncol(summer_expense_rep), rows = 1)
addStyle(wb_combined, "tables_unite", title_style_summer, rows = 1, cols = 1:ncol(summer_expense_rep), gridExpand = TRUE)
writeData(wb_combined, "tables_unite", summer_expense_rep, startRow = 2, startCol = 1, colNames = TRUE)
addStyle(wb_combined, "tables_unite", header_style_summer, rows = 2, cols = 1:ncol(summer_expense_rep), gridExpand = TRUE)

# add a blank space between the tables (2 columns of empty cells)
num_rows_summer <- nrow(summer_expense_rep)
writeData(wb_combined, "tables_unite", matrix("", nrow = num_rows_summer + 2, ncol = 2), startRow = 1, startCol = ncol(summer_expense_rep) + 1, colNames = FALSE)

# write the Analysis Table with title and formatting
writeData(wb_combined, "tables_unite", "Analysis Table", startRow = 1, startCol = ncol(summer_expense_rep) + 3)
mergeCells(wb_combined, "tables_unite", cols = (ncol(summer_expense_rep) + 3):(ncol(summer_expense_rep) + 2 + ncol(analysis_tbl)), rows = 1)
addStyle(wb_combined, "tables_unite", title_style_analysis, rows = 1, cols = (ncol(summer_expense_rep) + 3):(ncol(summer_expense_rep) + 2 + ncol(analysis_tbl)), gridExpand = TRUE)
writeData(wb_combined, "tables_unite", analysis_tbl, startRow = 2, startCol = ncol(summer_expense_rep) + 3, colNames = TRUE)
addStyle(wb_combined, "tables_unite", header_style_analysis, rows = 2, cols = (ncol(summer_expense_rep) + 3):(ncol(summer_expense_rep) + 2 + ncol(analysis_tbl)), gridExpand = TRUE)

# save the final workbook
# the file will automatically assign a date to the workbook
# add a space in the file output as well to ensure the data is not all bundled up

saveWorkbook(wb_combined, paste0("FY ", fiscal_year, " Budget and Actual as of ", format(current_date, "%m-%d-%Y"), ".xlsx"), overwrite = TRUE)
```


#### YOU DID IT ####
Congratulations! You did a wonderful job coding! 
If this is your first real coding r markdown file, then I must say you choose a tricky assignment. 
Go ahead and call yourself a Rstudio programmer you data analyst!
#### YOU DID IT ####


