---
title: "sript"
author: "Selenea Gibson and Jason Howard"
date: "2025-01-23"
output: html_document
---

# Project Overview
This markdown file will use the DHCD Finance dept data to calculate the columns that are provided from the `754 SFSP Expense Report`. There will be a series of plots to make the tables at the bottom of the report more appealing for what is being shown. This will be done through visuals 

```{r libraries, warning=FALSE, message=TRUE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
```

## Reading in the data 
Reading in the files for the data that will be used for creating the plots that will represent the data on the bottom of the `754 SFSP Expense Report`.

```{r readingdata}
# reading in the dataset for the summer_expense_report
summer_exp <- read.xlsx(here("FY 2024 Budget and Actual as of 12_31_2024.xlsx"))

# set the first row as headers and remove it from the data
summer_exp <- setNames(summer_exp[-1, ], as.character(unlist(summer_exp[1, ])))
```

## Creating plots
Here I want to create plots that relate to the data that we see in the table. These will be used 

```{r}

```







































































