---
title: "SRV0745 Housing Code Enforcement Plots Deputy Eric Booker"
author: "Selenea Gibson and Jason Howard"
date: "2025-01-21"
output: html_document
---

## Project Overview
This markdown file will use the DHCD Finance dept documents that are for making plots to help showcase the findings from the `Gl_and_Budget_vs_Actuals.rmd` file. This is part of a series of rmd files that will allow you to make plots for the Deputy by cost center.

*Note: This will make plots for `SRV0745 Housing Code Enforcement` only and will be for Deputy Eric Booker*

### Libraries

```{r libraries, warning=TRUE, message=FALSE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(ggplot2)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
```

## Reading in the data 
Reading in the files for the data that will be used. Put new data files in the "Data" folder and replace the file names below with the names of the new files.

```{r data}
# read in the data
# change the name of this file to represent the newest file downloaded
budg_vs_act_file_name <- "Budget_vs_Actuals_-_BBMR YTD 2024-12-19.xlsx"

# the first row of the data file is mostly empty except for a few important dates. 
first_row_of_data <- readxl::read_xlsx(here("Data", budg_vs_act_file_name)) |> 
  names() 

# read those data and do some formatting to clean up the data
name_prefixes <- tibble(first_row_of_data) |> 
  mutate(prefixes = ifelse(str_detect(first_row_of_data, "\\.\\."), 
                          NA_character_, 
                          first_row_of_data)) |> 
fill(prefixes, .direction = "down") |> 
  pull(prefixes)

# read the data file again but skip the first line
budg_vs_actuals <- readxl::read_xlsx(here("Data", budg_vs_act_file_name),
                  skip = 1)

#dropping the total budget row as this is not needed and will mess up the data
budg_vs_actuals <- budg_vs_actuals |> 
  filter(`Cost Center` != "Total")

# the names in the file repeat themselves, but this combines the dates with the repetative names
for(i in seq_along(name_prefixes)){
  plain_name <- gsub("\\.\\..*","", names(budg_vs_actuals)[i]) 
  
  names(budg_vs_actuals)[i] <- 
    ifelse(is.na(name_prefixes[i]) & str_length(plain_name) > 0,
           plain_name,
           paste(plain_name,name_prefixes[i], sep = "_"))
}

 names(budg_vs_actuals)
```

# Plots for Deputy Eric Booker
Making some plots that show different perspectives with the `budg_vs_actuals` df. These three plots will look at the following:
*Service* 
  1.`SRV0745 Housing Code Enforcement`
*Cost Center* 
  1.`CCA000326 Housing Code Property Maintenance Code Enforcement`
*Total Spending by Month* 

## Negative Spending by Cost Center
The plot below will show negative spending for service types & Cost Center vs. total spent.
*Note: If the plot has no values then there was no negative spending to be found for any of the cost centers*

```{r totspentnegplot}
# filter the dataset for negative spending values
hce_nh_negative_spending <- budg_vs_actuals |>
  filter(
    Service == "SRV0745 Housing Code Enforcement" &
    `Cost Center` %in% c("CCA000326 Housing Code Property Maintenance Code Enforcement") &
    `Total Spent` < 0
  )

# plotting data below for different perspectives
hce_nh_tot_neg_ccenter <- ggplot(hce_nh_negative_spending, aes(x = reorder(`Cost Center`, `Total Spent`), y = `Total Spent`)) +
  geom_col(fill = "red", width = 0.5) +  # Adjust bar width for better aesthetics
  theme_minimal() +  # Minimalistic theme
  scale_y_reverse(labels = scales::dollar) +  # Reverse y-axis and format with dollar labels
  labs(
    title = "Total Negative Spending by Cost Center",
    subtitle = "Service: SRV0745 Housing Code Enforcement",
    x = "Cost Center",
    y = "Total Spent (USD)",
    caption = "Data Source: BBMR Data"  # Add data source
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 22),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 22),  # Adjust y-axis text size
    axis.title.x = element_text(size = 24, face = "bold"),  # Emphasize x-axis title
    axis.title.y = element_text(size = 24, face = "bold"),  # Emphasize y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Center-align and enlarge title
    plot.subtitle = element_text(size = 24, hjust = 0.5, face = "italic"),  # Style subtitle
    plot.caption = element_text(size = 20, face = "italic"),  # Style caption
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA)  # Set plot background to white
  )

# display the plot
print(hce_nh_tot_neg_ccenter)

# saving the plot
ggsave("hce_nh_tot_neg_ccenter_deputy_eric.pdf", 
       width = 30, height = 20)
```

## Negative spending by Spend Category
This plot below will show the spending drilled down further by the Spend Category to highlight the areas where there is a negative cost.
*Note: If the plot has no values then there was no negative spending to be found for any of the cost centers*

```{r spendcat}
# plotting data below for different perspectives
hce_nh_tot_neg_spcat <- ggplot(hce_nh_negative_spending, aes(x = reorder(`Spend Category`, `Total Spent`), y = `Total Spent`)) +
  geom_col(fill = "red", width = 0.5) +  # Adjust bar width for better aesthetics
  theme_minimal() +
  scale_y_reverse(labels = scales::dollar) +  # Reverse y-axis and format with dollar labels
  labs(
    title = "Total Negative Spending by Spend Category",
    subtitle = "Service: SRV0745 Housing Code Enforcement",
    x = "Spend Category",
    y = "Total Spent (USD)",
    caption = "Data Source: BBMR Data"  # Add data source
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 22),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 22),  # Adjust y-axis text size
    axis.title.x = element_text(size = 24, face = "bold"),  # Emphasize x-axis title
    axis.title.y = element_text(size = 24, face = "bold"),  # Emphasize y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Center-align and enlarge title
    plot.subtitle = element_text(size = 24, hjust = 0.5, face = "italic"),  # Style subtitle
    plot.caption = element_text(size = 20, face = "italic"),  # Style caption
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA)  # Set plot background to white
  )

# display the plot
print(hce_nh_tot_neg_spcat)

# saving the plot
ggsave("hce_nh_tot_neg_spcat_deputy_eric.pdf",
       width = 30, height = 20)
```

## Actual Spending by month
The plot below will show spending for service types and the actual spent for a specific month.

```{r monthplot}
# filter the dataset for the actual spent
hce_nh_actual_spent <- budg_vs_actuals |>
  filter(
    Service == "SRV0745 Housing Code Enforcement" &
    `Cost Center` %in% c("CCA000326 Housing Code Property Maintenance Code Enforcement")
  )

# calculate the maximum spending for each cost center
hce_nh_max_spent_data <- hce_nh_actual_spent |> 
  group_by(`Cost Center`) |> 
  filter(`Actuals_2024-07` == max(`Actuals_2024-07`)) |> 
  ungroup()

# create a plot showing Actuals spent for July or a different plot
# by Cost center with positive (blue)
hce_nh_july_spending_plot <- ggplot(hce_nh_max_spent_data, aes(x = `Cost Center`, y = `Actuals_2024-07`)) +
  geom_col(fill = "blue", width = 0.5) +  # Set all bars to blue
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar) +  # Format y-axis to show dollar values
  labs(
    title = "Total Spending for Each Cost Center in July 2024",  # Update month/year
    subtitle = "Service: SRV0745 Housing Code Enforcement",
    x = "Cost Center",
    y = "July Actuals ($)",
    caption = "Data Source: BBMR Data"  # Add data source
  ) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 22),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 22),  # Adjust y-axis text size
    axis.title.x = element_text(size = 24, face = "bold"),  # Emphasize x-axis title
    axis.title.y = element_text(size = 24, face = "bold"),  # Emphasize y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Center-align and enlarge title
    plot.subtitle = element_text(size = 24, hjust = 0.5, face = "italic"),  # Style subtitle
    plot.caption = element_text(size = 20, face = "italic"),  # Style caption
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA),  # Set plot background to white
    legend.title = element_text(size = 24, face = "bold"),  # Bigger legend title
    legend.text = element_text(size = 22),  # Bigger legend text
    legend.key.size = unit(1.5, "lines")  # Increase size of the legend key (the color box)
  )

# Display the plot
print(hce_nh_july_spending_plot)

# saving the plot
ggsave("hce_nh_july_spending_deputy_eric.pdf", 
       width = 45, height = 30)
```
