---
title: "SRV0750 Housing Rehabilitation Services Plots Deputy Nicole Hart"
author: "Selenea Gibson and Jason Howard"
date: "2025-01-21"
output: html_document
---

## Project Overview
This markdown file will use the DHCD Finance dept documents that are for making plots to help showcase the findings from the `Gl_and_Budget_vs_Actuals.rmd` file. This is part of a series of rmd files that will allow you to make plots for the Deputy by cost center.

*Note: This will make plots for `SRV0750 Housing Rehabilitation Services` only and will be for Deputy Nicole Hart*

### Libraries

```{r libraries, warning=TRUE, message=FALSE}
# libraries for data analysis
library(dplyr)
library(here)
library(tidyverse)
library(scales)
library(ggplot2)
library(DT)
library(openxlsx)
library(stringr)
library(lubridate)
```

## Reading in the data 
Reading in the files for the data that will be used. Put new data files in the "Data" folder and replace the file names below with the names of the new files.

```{r data}
# Function to get the latest file in the data folder based on a pattern
get_latest_file <- function(pattern) {
  files <- list.files(here("Data"), pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(NULL)  # Handle case when no file matches
  latest_file <- files[order(file.info(files)$mtime, decreasing = TRUE)][1]
  return(latest_file)
}

# Get the latest Budget vs Actuals file
budg_vs_act_file <- get_latest_file("Budget_vs_Actuals_-_BBMR YTD .*\\.xlsx$")

# the first row of the data file is mostly empty except for a few important dates. 
# read those data and do some formatting to clean up the data
first_row_of_data <- readxl::read_xlsx(budg_vs_act_file, col_names = T) |> names()

# Extract prefixes for column names
name_prefixes <- tibble(first_row_of_data) |> 
  mutate(prefixes = ifelse(str_detect(first_row_of_data, "\\.\\."), 
                           NA_character_, 
                           first_row_of_data)) |> 
  fill(prefixes, .direction = "down") |> 
  mutate(prefixes = ifelse(is.na(prefixes), NA_character_, paste0(prefixes, "-01"))) |> 
  pull(prefixes)

# Read the full dataset, skipping the first row
budg_vs_actuals <- readxl::read_xlsx(budg_vs_act_file, skip = 1)

# remove the total budget row as this is not needed 
# and will mess up the data. 
budg_vs_actuals <- budg_vs_actuals |> 
  filter(`Cost Center` != "Total") 

# the names in the file repeat themselves, but this combines the dates with the repetitive names
for(i in seq_along(name_prefixes)){
  plain_name <- gsub("\\.\\..*","", names(budg_vs_actuals)[i]) 
  
  names(budg_vs_actuals)[i] <- 
    ifelse(is.na(name_prefixes[i]) & str_length(plain_name) > 0,
           plain_name,
           paste(plain_name,name_prefixes[i], sep = "_"))
}

# remove commitments and obligations since we dont need them
budg_vs_actuals <- budg_vs_actuals |> 
  select(-matches("^Commitments|^Obligations"))

# remove columns that have nulls for all values 
# replace "blank" with an actual 'null' value
budg_vs_actuals <- budg_vs_actuals |> 
  select_if(~sum(!is.na(.)) > 0) |> 
  mutate(Grant = ifelse(Grant == "(Blank)", NA_character_, Grant))

# looking at the new names 
names(budg_vs_actuals)
```

# Plots for Deputy Nicole Hart
Making some plots that show different perspectives with the `budg_vs_actuals` df. These three plots will look at the following:
*Service* 
  1.`SRV0750 Housing Rehabilitation Services`
*Cost Center* 
  1.`CCA000339 Housing Rehabilitation Loans`
  2.`CCA000340 Lead Abatement (HCD)`
  3.`CCA001177 Housing Rehabilitation Services Unallocated Appropriation`
  
*Total Spending by Month* 

## Negative Spending by Cost Center
The plot below will show negative spending for service types & Cost Center vs. total spent.
*Note: If the plot has no values then there was no negative spending to be found for any of the cost centers*

```{r totspentnegplot}
# filter the dataset for negative spending values
hrs_nh_negative_spending <- budg_vs_actuals |>
  filter(
    Service == "SRV0750 Housing Rehabilitation Services" &
    `Cost Center` %in% c("CCA000339 Housing Rehabilitation Loans", 
                         "CCA000340 Lead Abatement (HCD)",
                         "CCA001177 Housing Rehabilitation Services Unallocated Appropriation") &
    `Total Spent` < 0
  )

# plotting data below for different perspectives
hrs_nh_tot_neg_ccenter <- ggplot(hrs_nh_negative_spending, aes(x = reorder(`Cost Center`, `Total Spent`), y = `Total Spent`)) +
  geom_col(fill = "red", width = 0.5) +  # Adjust bar width for better aesthetics
  theme_minimal() +  # Minimalistic theme
  scale_y_reverse(labels = scales::dollar) +  # Reverse y-axis and format with dollar labels
  labs(
    title = "Total Negative Spending by Cost Center",
    subtitle = "Service: SRV0750 Housing Rehabilitation Services",
    x = "Cost Center",
    y = "Total Spent (USD)",
    caption = "Data Source: BBMR Data"  # Add data source
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 22),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 22),  # Adjust y-axis text size
    axis.title.x = element_text(size = 24, face = "bold"),  # Emphasize x-axis title
    axis.title.y = element_text(size = 24, face = "bold"),  # Emphasize y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Center-align and enlarge title
    plot.subtitle = element_text(size = 24, hjust = 0.5, face = "italic"),  # Style subtitle
    plot.caption = element_text(size = 20, face = "italic"),  # Style caption
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA)  # Set plot background to white
  )

# display the plot
print(hrs_nh_tot_neg_ccenter)

# saving the plot
ggsave("hrs_nh_tot_neg_ccenter_deputy_Hart.pdf",
       width = 30, height = 20)
```

## Negative spending by Spend Category
This plot below will show the spending drilled down further by the Spend Category to highlight the areas where there is a negative cost.
*Note: If the plot has no values then there was no negative spending to be found for any of the cost centers*

```{r spendcat}
# plotting data below for different perspectives
hrs_nh_tot_neg_spcat <- ggplot(hrs_nh_negative_spending, aes(x = reorder(`Spend Category`, `Total Spent`), y = `Total Spent`)) +
  geom_col(fill = "red", width = 0.5) +  # Adjust bar width for better aesthetics
  theme_minimal() +
  scale_y_reverse(labels = scales::dollar) +  # Reverse y-axis and format with dollar labels
  labs(
    title = "Total Negative Spending by Spend Category",
    subtitle = "Service: SRV0750 Housing Rehabilitation Services",
    x = "Spend Category",
    y = "Total Spent (USD)",
    caption = "Data Source: BBMR Data"  # Add data source
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 22),  # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 22),  # Adjust y-axis text size
    axis.title.x = element_text(size = 24, face = "bold"),  # Emphasize x-axis title
    axis.title.y = element_text(size = 24, face = "bold"),  # Emphasize y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Center-align and enlarge title
    plot.subtitle = element_text(size = 24, hjust = 0.5, face = "italic"),  # Style subtitle
    plot.caption = element_text(size = 20, face = "italic"),  # Style caption
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA)  # Set plot background to white
  )

# display the plot
print(hrs_nh_tot_neg_spcat)

# saving the plot
ggsave("hrs_nh_tot_neg_spcat_deputy_Hart.pdf", 
       width = 30, height = 20)
```

## Actual Spending by month
The plot below will show spending for service types and the actual spent for a specific month.

```{r monthplot, warning=FALSE, message=FALSE}
# filter the dataset for the actual spent
hrs_nh_actual_spent <- budg_vs_actuals |>
  filter(
    Service == "SRV0750 Housing Rehabilitation Services" &
    `Cost Center` %in% c("CCA000339 Housing Rehabilitation Loans", 
                         "CCA000340 Lead Abatement (HCD)",
                         "CCA001177 Housing Rehabilitation Services Unallocated Appropriation")
  )

# calculate the maximum spending for each cost center
hrs_nh_actual_spent_long <- hrs_nh_actual_spent |> 
  pivot_longer(
    cols = starts_with("Actuals_"),  
    names_to = "Month", 
    values_to = "Actual_Spent"
  ) |> 
  mutate(Month = str_remove(Month, "Actuals_")) |>  # Remove "Actuals_" prefix
  mutate(Month = gsub("\\.", "-", Month)) |>  # Replace dot with hyphen
  mutate(Month = as.Date(paste0(Month, "-01"), format = "%Y-%m-%d"))  # Convert to Date format with a default day (01)


# Ensure the Month column has valid entries and remove any rows with NA values
hrs_nh_actual_spent_long <- hrs_nh_actual_spent_long |> 
  filter(!is.na(Month))

# Now, create a sequence of months from the minimum to the current month
full_months <- tibble(Month = seq(min(hrs_nh_actual_spent_long$Month), 
                                  today(), 
                                  by = "month"))


# Merge the full month sequence with the actual data and replace NAs with 0
hrs_nh_actual_spent_long <- full_months |> 
  left_join(hrs_nh_actual_spent_long, by = "Month") |> 
  mutate(Actual_Spent = replace_na(Actual_Spent, 0))  # Fill missing values with 0


# Create the plot with y-axis limits adjusted
hrs_nh_actual_spent_plot <- ggplot(hrs_nh_actual_spent_long, aes(x = Month, y = `Actual_Spent`, fill = `Cost Center`)) +
  geom_col(position = "dodge", width = 30) +  # Dodge bars by Cost Center
  theme_minimal() +
  scale_y_continuous(
    labels = scales::dollar,  # Format y-axis as dollars
    limits = c(0, 300000)  # Set y-axis limits from 0 to 500,000
  ) + 
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +  # Show months clearly
  coord_flip() +
  labs(
    title = "Total Spending for Each Cost Center Over Time",
    subtitle = "Service: SRV0750 Housing Rehabilitation Services",
    x = "Month",
    y = "Monthly Spend ($)",
    caption = "Data Source: BBMR Data",
    fill = "Cost Center"  # Rename the legend title
  ) + 
  scale_fill_discrete(na.translate = FALSE) +  # Exclude NA from the legend, but include in plot
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 24, face = "bold"),  # Rotate and bold x-axis labels for readability
    axis.text.y = element_text(size = 24, face = "bold"),  # Bold y-axis text
    axis.title.x = element_text(size = 26, face = "bold"),  
    axis.title.y = element_text(size = 26, face = "bold"),  
    plot.title = element_text(size = 30, face = "bold", hjust = 0.5),  
    plot.subtitle = element_text(size = 28, hjust = 0.5, face = "italic"),  
    plot.caption = element_text(size = 16, face = "italic"),  
    panel.grid.major = element_line(color = "gray90"),  # Lighter grid for subtle effect
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    plot.background = element_rect(fill = "white", color = NA),  # Set plot background to white
    legend.title = element_text(size = 26, face = "bold"),  # Adjust legend title size
    legend.text = element_text(size = 24),  # Adjust legend text size
    legend.key.size = unit(3, "lines")  # Increase size of the legend key
  )

# Display the plot
print(hrs_nh_actual_spent_plot)

# saving the plot
ggsave("hrs_nh_monthly_spending_deputy_Hart.pdf",
       width = 45, height = 30)
```

